---
title: "DE strategy robustness"
output: html_document
---

### Aim of this analysis

Test to what extent our DE strategy is robust. I.e. Using all "supergene" expression values when fitting the model followed by contrast tests. 

In some species, i.e. StLa and NaSt there are only a single sample for D0 and D1, so that the short term cold reponse effect is based on the difference between two samples with no replicates. This is deemed insufficient to determine statistically significant DE. MeNu1 on the other hand has plenty of replicates and would be deemed to be sufficient.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = 'svg')
```


```{r loadLibraries, echo=FALSE, message=FALSE}
library(DESeq2)
# library(BiocParallel)
# library(gplots)
library(tidyverse)
library(plotly)
library(ape) # plot.phylo
library(phangorn) # NJ

# BPparam <- MulticoreParam(workers = 20)
```

```{r loadData, echo=FALSE}
expr <- readRDS("~/GRewd/pipeline/data/superExprMatrix.RDS") # from superGeneModel.Rmd
vst <- readRDS("~/GRewd/pipeline/data/VSTnew.RDS")

```

```{r colDataAndNormalize }
spcs <- c("BrDi","HoVu","MeNu1","StLa","NaSt")
spcs <- setNames(spcs,spcs)

TtoF <- c(`T-1`="ramp0",T0="peak0",T1="peak1",T3="ramp1",T4="ramp1")

colData <- as.data.frame(stringr::str_split_fixed(colnames(expr),"\\.",4)[ ,1:3])
names(colData) <- c("spcPop","timePoint","mix")
colData$spc <-  as.factor(stringr::str_extract(colData$spcPop,paste(spcs,collapse="|")))

colData$sf <- as.factor(paste0(colData$spc,TtoF[as.character(colData$timePoint)]))


dds <- DESeqDataSetFromMatrix(countData = round(expr),
                              colData = colData,
                              design = formula(~ sf ))
  
# dds <- DESeq(dds,parallel = T,BPPARAM = BPparam)

# just need to normalize
dds <- estimateSizeFactors(dds)
```

### Same variance across species and timepoints

To compare the replicate variance I used the individual samples for which there are triplicates in a selection of species and timepoints. For each gene the variance of the log transformed normalized read-count is calculated over the three (or four) replicates. For each pair of time/spc sample points with replicates, the spearman correlation between the variances are calculated using the around 3000 genes that are highly expressed in both samples points (log2(mean(normCount)) > 8). Need to use the highly expressed genes as the variance correlates strongly with the mean at low expression levels, while at high expression levels the variance is dominated by the biological variance.

The plot shows the resulting spearman coefficients divided into comparisons within species, within same time point, or both different timepoint and species. There is no observable difference when comparing within species or within time point, but seemingly lower correlation when both species and time point are different.

Apparently the variance is neither the same within species nor between species, but both factors have about the same effect. Therefore it makes as much same to assume common variance accross species as it does to assume common variance between timepoints. 

```{r}

# get individual samples
x <- counts(dds,normalized = T)[ ,colData$mix=="ind"]

# get spc+timepoint factor
spcT <- sub("(.*)\\.(ind|mix)..","\\1",colnames(x))

table(spcT)

# means for each spc+timepoint
do.call(cbind,tapply(colnames(x),spcT,function(cols){
  log2(apply(x[,cols],1,mean))
})) -> means

# vars for each spc+timepoint
do.call(cbind,tapply(colnames(x),spcT,function(cols){
  apply(log2(x[,cols]),1,var)
})) -> vars

minMeans <- 8

sapply(colnames(vars),function(i1){
  sapply(colnames(vars),function(i2){
    # only include highly expressed and remove missing
    idx <- !is.na(vars[ ,i1]) & !is.nan(vars[ ,i1]) & means[ ,i1] > minMeans &
       !is.na(vars[ ,i2]) & !is.nan(vars[ ,i2]) & means[ ,i2] > minMeans

    cor(vars[idx, i1],vars[idx, i2],method = "spearman")
  })
}) -> varCor
# diag(varCor) <- NA
# heatmap(varCor)

as_data_frame(varCor) %>% 
  mutate( spcT1=rownames(varCor)) %>% 
  gather( key="spcT2",value = "varCor", -spcT1) %>% 
  filter( spcT1 < spcT2  ) %>% 
  mutate( sameSpc = sub("(.*)\\.(.*)","\\1",spcT1) == sub("(.*)\\.(.*)","\\1",spcT2),
          sameT = sub("(.*)\\.(.*)","\\2",spcT1) == sub("(.*)\\.(.*)","\\2",spcT2),
          isSame = as.factor(ifelse(sameT,"TimePoint",ifelse(sameSpc,"Species","Neither"))),
          compared = paste(spcT1,spcT2)) %>% 
  ggplot( aes( y=varCor, x = isSame, label=compared)) + geom_boxplot() + geom_point() -> p


ggplotly(p)

```


```{r eval=FALSE}
# exclude the MeNu1 peak samples that are not pooled
excludeSamples <- colData$sf %in% c("MeNu1peak0","MeNu1peak1") & colData$mix=="ind"

dds <- DESeqDataSetFromMatrix(countData = round(expr[ ,!excludeSamples]),
                              colData = colData[!excludeSamples, ],
                              design = formula(~ sf ))
  
dds <- DESeq(dds,parallel = T,BPPARAM = BPparam)

ddsMeNuPooled <- dds
```


```{r eval=FALSE}
# exclude all but a single of the MeNu1 peak samples that are not pooled

lapply(1:3, function(i){
  excludeSamples <- colData$sf %in% c("MeNu1peak0","MeNu1peak1")
  includeSamples <- c(which(colData$sf == "MeNu1peak0" & colData$mix=="ind")[i],
                      which(colData$sf == "MeNu1peak1" & colData$mix=="ind")[i])
  idx <- sort(c(which(!excludeSamples),includeSamples))
  
  dds <- DESeqDataSetFromMatrix(countData = round(expr[ ,idx]),
                              colData = colData[idx, ],
                              design = formula(~ sf ))
  
  DESeq(dds,parallel = T,BPPARAM = BPparam)
}) -> ddsMeNuSingle

```


```{r eval=FALSE}
includeSamples <- colData$sf %in% c("MeNu1peak0","MeNu1peak1") & colData$mix=="ind"

dds <- DESeqDataSetFromMatrix(countData = round(expr[ ,includeSamples]),
                              colData = colData[includeSamples, ],
                              design = formula(~ sf ))
  
dds <- DESeq(dds,parallel = T,BPPARAM = BPparam)

ddsMeNuSimple <- dds



```

```{r eval=FALSE}
# collect results

# short term all species
lapply(spcs, function(spc){
  results(ddsFull, contrast=c("sf",paste0(spc,"peak1"),paste0(spc,"peak0")))  
}) -> resFull

# variants of short term MeNu

resMeNu <-
  list(
    pooled = results(ddsMeNuPooled, contrast=c("sf","MeNu1peak1","MeNu1peak0")),
    single1 = results(ddsMeNuSingle[[1]], contrast=c("sf","MeNu1peak1","MeNu1peak0")),
    single2 = results(ddsMeNuSingle[[2]], contrast=c("sf","MeNu1peak1","MeNu1peak0")),
    single3 = results(ddsMeNuSingle[[3]], contrast=c("sf","MeNu1peak1","MeNu1peak0")),
    classic = results(ddsMeNuSimple, contrast=c("sf","MeNu1peak1","MeNu1peak0"))
  )

resMeNu$full <- resFull$MeNu1

```

```{r eval=FALSE}
sigSTI <- lapply(resFull, with, { ifelse(is.na(padj), F, padj < 0.05 & log2FoldChange > 1) })
sigSTIMeNu <- lapply(resMeNu, with, { ifelse(is.na(padj), F, padj < 0.05 & log2FoldChange > 1) })

```

```{r eval=FALSE}
venn(lapply(sigSTI,which))
```

```{r eval=FALSE}

plotOverlap(sigSTIMeNu[-6],FUN = oddsRatioZ,xlab = "Z-score")
plotOverlap(sigSTI,FUN = oddsRatio,xlab = "odds ratio")
plotOverlap(sigSTI,FUN = oddsRatioFisher,xlab = "Z-score")
plotOverlap(c(sigSTIMeNu[-6],sigSTI[-3]),FUN = oddsRatioZ,xlab = "Z-score")


```


```{r eval=FALSE}
myOuter <- function(X,Y=X,FUN){
  apply(X,2,function(x){
    apply(Y,2,function(y){
      FUN(x,y)
    })
  })
}

nBoth <- function(m){
  myOuter(m,FUN=function(x,y){ sum(x & y) })
}

oddsRatio <- function(m){
  n11 <- myOuter(m,FUN=function(x,y){ sum( x &  y) })
  n00 <- myOuter(m,FUN=function(x,y){ sum(!x & !y) })
  n10 <- myOuter(m,FUN=function(x,y){ sum( x & !y) })
  n01 <- myOuter(m,FUN=function(x,y){ sum(!x &  y) })

  (n00*n11)/(n10*n01)
}

oddsRatioSE <- function(m){
  n11 <- myOuter(m,FUN=function(x,y){ sum( x &  y) })
  n00 <- myOuter(m,FUN=function(x,y){ sum(!x & !y) })
  n10 <- myOuter(m,FUN=function(x,y){ sum( x & !y) })
  n01 <- myOuter(m,FUN=function(x,y){ sum(!x &  y) })

  sqrt(1/n00 + 1/n11 +1/n10 + 1/n01)
}

oddsRatioZ <- function(m){
  log(oddsRatio(m))/oddsRatioSE(m)
}

pnorm(6.342536,lower.tail = F)

oddsRatioFisher <- function(m){
  retVal <- matrix( 0, ncol=ncol(m), nrow=ncol(m), dimnames=list(colnames(m),colnames(m)))
  
  retVal[lower.tri(retVal)] <-
    list(
      n11 = myOuter(m,FUN=function(x,y){ sum( x &  y) }),
      n10 = myOuter(m,FUN=function(x,y){ sum( x & !y) }),
      n01 = myOuter(m,FUN=function(x,y){ sum(!x &  y) }),
      n00 = myOuter(m,FUN=function(x,y){ sum(!x & !y) })
    ) %>% 
    map( ~ .x[lower.tri(.x)]) %>% 
    pmap( c ) %>% 
    map( matrix, nrow=2) %>% 
    map( fisher.test, alternative="greater") %>% 
    map_dbl( ~ .x$estimate)
  
  return(retVal)
}


pmap(list(n11,n10,n01,n00),c)

x <- fisher.test(matrix(c(136,308,533,7656),nrow=2),alternative="greater")
x$estimate

plotOverlap <- function(sigList, FUN=oddsRatio, xlab="Odds ratio"){
  m <- do.call(cbind,sigList)
  plotTriMat(labelMat=nBoth(m),valueMat = FUN(m), xlab=xlab)
}

plotTriMat <- function(labelMat,valueMat,xlab){
  valRange <- range(pretty(valueMat[lower.tri(valueMat)]))
#   valColors <- colorRampPalette(c("#FFEDA0","#FEB24C","#F03B20"))(100)
  valColors <- colorRampPalette(RColorBrewer::brewer.pal(7,"YlGnBu")[1:5])(100)
  valBreaks <- seq(valRange[1],valRange[2],length.out = 101)
  
  colorMat = matrix( cut(valueMat, labels=valColors, breaks=valBreaks), nrow=nrow(valueMat))
  
  n <- ncol(valueMat)
  
  layout(mat = matrix(1:2,ncol=1),heights = c(3,1))
  par(mar=c(0,1,2,1)+0.1)
  plot.new()
  plot.window( xlim=c(-5,n), ylim=c(-0.5,2*n-1.5), asp=1 )
  x=c(-1,0,1,0);y=c(0,1,0,-1)
  for(i in 1:n){
    for(j in 1:i){
      cellCenterX <- i-j
      cellCenterY <- 2*n - (i+j)
      if(i != j){
        polygon(x = cellCenterX+x, cellCenterY+y,border = "black",col = colorMat[i,j])
        text(x = cellCenterX, y = cellCenterY-0.4, labels = signif(valueMat[i,j],digits = 3),cex=0.8)
      }
      text(x = cellCenterX, y = cellCenterY, labels = labelMat[i,j])
    }
  }

  # plot species names in italic
  for(i in 1:n){
    text(x = -1, y = 2*(n-i),adj = c(1,0.5), 
         labels = substitute(italic(s),list(s=colnames(labelMat)[i])))
  }
  
  title(main="Pairwise intersect of DEGs")

  par(mar=c(4,6,1,6)+0.1)
  image(x = valBreaks,z = matrix(valBreaks,ncol=1),col=valColors,
        yaxt = "n", xlab=xlab)

  par(mar=c(5, 4, 4, 2) + 0.1)
  layout(1)
}  
```

## Pooled sample variance test

Among the samples there are 10 species/timpoint combinations that have exactly 1 pooled and 3 individual replicates. These can be used to test the pooling strategy, i.e. whether the pooled samples have less variance than individual samples. By sorting the expression level of an indivual gene among the 3+1 replicates we would expect the pooled replicate to end up in the middle if it has lower variance. The figure below shows the distribution of ranks of the pooled replicate. It shows that 6 out of 10 behaves as expected, i.e. mostly rank 2-3, while 4 (Stipa T4, Nardus T4, BrDi T3 og Melica T1) seem to have higher or equal variance in the pooled samples.

```{r pooledVariance}
hasTbl <- 
  colData %>%
  group_by(spc, timePoint) %>%
  summarise(has3ind1mix = (sum(mix=="ind")==3 & sum(mix=="mix")==1)) %>%
  filter(has3ind1mix) %>%
  mutate(spcT=paste(spc,timePoint,sep="."))

includeSamples <-
  colData %>% transmute(spcT=paste(spc,timePoint,sep=".")) %>%
  .$spcT %in% hasTbl$spcT


x <- counts(dds,normalized=T)[,includeSamples]


tapply(colnames(x),sub("(.*)\\.(mix|ind)..","\\1",colnames(x)),function(sampleID){
  y <- x[ ,sort(sampleID)]
  z <- t(apply(y,1,rank,ties.method="random"))
  table(z[ ,4])
}) -> tbls

do.call(rbind,tbls) %>% 
  as_data_frame() %>% 
  mutate(sampleID=names(tbls)) %>% 
  gather(c(1:4),key="rank",value = "count") %>% 
  mutate(rank=as.integer(rank)) %>% 
  ggplot( aes(x=rank,y=count,color=sampleID)) + 
  geom_line()


```

## sample clustering

Large points indicate pooled samples. In the first plot the pooled samples tend to have a bit shorter branches (indicating less noise) and they tend to bunch between the individual replicates. The Stipa T4(W9) and Nardus T4(W9) samples that seemed to have higher variance seem to fall outside the bunch of individual samples. In the second plot the individual replicates are averaged for each spcs/timepoint. Here the averaged individual replicates tend to have shorter branches than the pooled indicating that they have less variance, although this is slighlty biased by subtracting the means. The third and fourth plot have not subtracted the means and in the last plot the mean of the individuals are not shorter than the pooled.

All in all it seems that average of individual replicates makes them closer to the pooled samples in most cases, indicating that the pooled samples has less of the within species variance.

```{r sampleTree}


# Species names to show in plots
spcID2spcName <- c( BrDi="B. distachyon", HoVu="H. vulgare", MeNu1="M. nutans",
                    StLa="S. lagascae", NaSt="N. stricta")

# define species colors to use in plots
spcColors <- c( NaSt = "#0049dbff",
                StLa = "#ff0000ff",
                MeNu1 = "#cd28e3ff",
                BrDi = "#00c500ff",
                HoVu = "#ff7e00ff")


sampleDesc <- read.table(text=colnames(vst),sep=".",col.names=c("spc","timepoint","pooled","rep"))
sampleDesc$spcColor <- spcColors[spcID2spcName[sampleDesc$spc]]
sampleDesc$spcName <- spcID2spcName[sampleDesc$spc]
sampleDesc$timePch <- c("-",1,2,3,4)[sampleDesc$timepoint]
sampleDesc$timeColor <- RColorBrewer::brewer.pal(11,"Spectral")[c(1,4,7,9,11)][sampleDesc$timepoint]

filterLowExpr <- apply(vst>3,1,all)

Reduce(cbind, sapply(unique(sampleDesc$spc),function(spc){
  isSpc <- sampleDesc$spc==spc
  # get mean per timepoint
  sapply(as.character(unique(sampleDesc$timepoint[isSpc])),function(timePoint){
    rowMeans(vst[ ,isSpc & (sampleDesc$timepoint==timePoint), drop=F])
  }) -> tpMeans
  #   # BrDi doesn't have T4.. does it help to use T3 as T4?
  #   if(!("T4" %in% colnames(tpMeans))){
  #     tpMeans <- cbind(tpMeans,T4=tpMeans[ ,"T3"]) # copy T3 where T4 is missing in Brachy
  #   }
  
  # For each species, subtract mean of timepoint means
  vst[ ,isSpc] - rowMeans(tpMeans)
})) -> x


meanReps <- function(z){
  # calculate mean of individual replicates
  tapply(colnames(z),sub(".[0-9]$","",colnames(z)), function(sampleID){
    rowMeans(z[,sampleID,drop=F])
  }) -> y
  
  do.call(cbind,y)
}

plotSampleTree <- function(z){
  dm <- dist(t(z),method = "manhattan")
  tree <- NJ(dm)
  
  plot.phylo(tree, type="u", show.tip.label = F,
             tip.color=spcColors[match(substr(tree$tip.label,1,1),substr(names(spcColors),1,1))],
             cex = 1)
  
  newTimepoints <- c(`T-1`="W0",T0="D0",T1="D1",T3="W4",T4="W9")
  tiplabels( frame = "none", text = newTimepoints[sub(".*(T-?[0134]).*","\\1",tree$tip.label,perl=T)],
             col=spcColors[match(substr(tree$tip.label,1,1),substr(names(spcColors),1,1))])
  tiplabels( pch=20,cex=ifelse(grepl("mix",tree$tip.label),2,0.8))
  legend("bottomleft",legend = names(spcColors),text.col = spcColors, y.intersp = 0.8,bty = "n" )
}

plotSampleTree(x[filterLowExpr, ])
plotSampleTree(meanReps(x[filterLowExpr, ]))
plotSampleTree(vst[filterLowExpr, ])
plotSampleTree(meanReps(vst[filterLowExpr, ]))


```


