---
title: "GeneListTest"
output:
  html_document:
    toc: yes
---

```{r loadLibraries, echo=FALSE, message=FALSE, results='hide'}
library(DESeq2)
library(gplots)
library(stringr)
library(topGO)


source("~/GRewd/pipeline/R/orthoGrpTools.R")
```

```{r loadData, echo=FALSE}
orthoPath <- "/mnt/NOBACKUP/mariansc/share/orthos"

DEnewFile<-"~/GRewd/pipeline/data/DEnew.RDS"

# load DE based based on superGenes
DEnew <- readRDS(DEnewFile)

# grpIDs included the superGeneModel.Rmd:
allGrpIDs <- rownames(DEnew$BrDi$resRamp)

# convert DE results to matrix format
lapply( list( peak="resPeak", ramp="resRamp"), function( peakOrRamp ){
  lapply( list( pVal="pvalue", pAdj="padj", stat="stat",lfc="log2FoldChange", lfcSE="lfcSE"), function( stat ){
    m <- sapply(DEnew, function(DEspc){DEspc[[peakOrRamp]][[stat]]})
    rownames(m) <- rownames(DEnew[[1]][[peakOrRamp]])
    return(m)
  })
}) -> DEmat

# load groups
grpData <- readRDS(file.path(orthoPath,"splitGroups/goodTreeStats.rds"))
grps <- loadOrthoGrpsArray(file.path(orthoPath,"splitGroups/goodGroups.txt"))

###########
# load paml results
pamlResFile<-"~/GRewd/pipeline/data/pamlRes.RDS"

pamlRes <- readRDS(pamlResFile) 

# remove NA's in H4a (probably caused by trees where the core clade is on the root and can't be marked)
pamlRes$H4a <- pamlRes$H4a[!is.na(pamlRes$H4a$pVal), ]

# convert to matrix of p-values
pamlMat <- sapply(pamlRes, with, {pVal[match(allGrpIDs,grpID)]})



# load barley gene annotations..
annotFile <- "/mnt/NOBACKUP/mariansc/share/refGenomes/barley_HighConf_genes_MIPS_23Mar12_HumReadDesc.txt"
annotFileSource <- "ftp://ftpmips.helmholtz-muenchen.de/plants/barley/public_data/genes/barley_HighConf_genes_MIPS_23Mar12_HumReadDesc.txt"

if(!file.exists(annotFile)){
  download.file(annotFileSource, destfile=annotFile)
}

annotTbl <- read.table(annotFile, as.is = T,sep = "\t",header = F,skip = 3, quote="",
           col.names = c("seqID","BlastHit","AHRD.QualityCode","Description","Interpro.annot","V6"))

annotTbl$grpID <- seqIDtoGrpID(grps, annotTbl$seqID, "Hv_R")

apply(annotTbl[ ,1:5],2,function(annotCol){
  tapply(annotCol, annotTbl$grpID, paste, collapse=" & ")
}) -> grpAnnotTbl

# as data.frame
grpAnnotTbl <- as.data.frame(grpAnnotTbl,stringsAsFactors = F)
```

```{r loadGOannot, echo=FALSE}
GOannotFile <- "/mnt/NOBACKUP/mariansc/share/refGenomes/barley_HighConf_genes_MIPS_23Mar12_IPROScan_GOs.txt"

# read GO annotations
HvGOtbl <- read.table(GOannotFile,sep="\t", quote = "",
                      col.names=c("geneID","GO"), stringsAsFactors=F)
# convert to gene2GO list
HvGene2GO <- split(HvGOtbl$GO,HvGOtbl$geneID)

```

```{r defGOFunctions, echo=FALSE}

# make grp2GO list
HvGOtbl$grpID <- seqIDtoGrpID(grps, HvGOtbl$geneID, "Hv_R")
grpID2GO <- lapply( split( HvGOtbl$GO, HvGOtbl$grpID ), unique )


testGOgrps <- function(grpIDs, bkGrpIDs = NULL, ontology = "BP"){
  # get annotated background groups
  if( !is.null(bkGrpIDs) ){
    annBkGrpIDs <- intersect(names(grpID2GO),bkGrpIDs)
  } else {
    annBkGrpIDs <- names(grpID2GO)
  }
  
  allGenes <- factor(as.integer(annBkGrpIDs %in% grpIDs))
  names(allGenes) <- annBkGrpIDs
  
  sink("/dev/null") # silence the output from topGO
  
  GOdata <- new("topGOdata", ontology = ontology, allGenes = allGenes,
                annot = annFUN.gene2GO, gene2GO = grpID2GO)
  
  
  # run fisher test using classic algorithm
  classic.fisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")
  weight.fisher <- runTest(GOdata, algorithm = "weight", statistic = "fisher")
  
  sink() # stop sinking output
  
  GenTable(GOdata, classic = classic.fisher, weight=weight.fisher,
           orderBy = "weight", ranksOf="classic", topNodes = 20)
}
```


### Explanation of the AHRD quality code

AHRD’s quality-code consists of a four character string, where each character is either ‘*’ if the respective criteria is met or ‘-’ otherwise. Their meaning is explained in the following table:

Position | Criteria
-------- | --------
   1     | Bit score of the blast result is >50 and e-value is < e-10
   2     | Overlap of the blast result is >60%
   3     | Top token score of assigned HRD is >0.5
   4     | Gene ontology terms found in description line 

### Genes responding to cold in all pooids

Which genes are up-regulated in all our five pooids as response to cold?

Select genes with p.adj<0.05 and lfc>1 in all species. Ramp and Peak effect tested individually.


```{r upInall, echo=FALSE}
sigMat <- lapply(DEmat, with, {pAdj < 0.05 & lfc > 1})

venn(apply(sigMat$peak,2,which))
title(main="significant peak up",sub="p.adj<0.05 & lfc>1")

whichAllpeakUp <- which(apply(sigMat$peak,1,all))

heatmap(DEmat$peak$stat[whichAllpeakUp,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13),
        RowSideColors = ifelse(pamlMat[whichAllpeakUp,"H4a"]<0.05,"red","grey"),
        scale="none", Rowv = NA, Colv = NA, main="Peak up in all (H4a)")


DT::datatable(grpAnnotTbl[names(whichAllpeakUp), ])

venn(apply(sigMat$ramp,2,which))
title(main="significant ramp up",sub="p.adj<0.05 & lfc>1")

whichAllrampUp <- which(apply(sigMat$ramp,1,all))

heatmap(DEmat$ramp$stat[whichAllrampUp,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13),
        RowSideColors = ifelse(pamlMat[whichAllrampUp,"H4a"]<0.05,"red","grey"),
        scale="none", Rowv = NA, Colv = NA, main="Ramp up in all (H4a)")

DT::datatable(grpAnnotTbl[names(whichAllrampUp), ])

```

### Only up in core

```{r simpleCoreBasal, echo=FALSE}
whichCorePeakUp <- which(apply(sigMat$peak[ ,1:2],1,all) & apply(!sigMat$peak[ ,3:5],1,all))

heatmap(DEmat$peak$stat[whichCorePeakUp,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13),
        RowSideColors = ifelse(pamlMat[whichCorePeakUp,"H4b"]<0.05,"red","grey"),      
        scale="none", Rowv = NA, Colv = NA, main="Peak up in core (H4b)")

DT::datatable(grpAnnotTbl[names(whichCorePeakUp), ])


whichCoreRampUp <- which(apply(sigMat$ramp[ ,1:2],1,all) & apply(!sigMat$ramp[ ,3:5],1,all))

heatmap(DEmat$ramp$stat[whichCoreRampUp,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13), Rowv = NA,
        RowSideColors = ifelse(pamlMat[whichCoreRampUp,"H4b"]<0.05,"red","grey"),      
        scale="none", Colv = NA, main="Ramp up in core (H4b)")

DT::datatable(grpAnnotTbl[names(whichCoreRampUp), ])
```

### More up in core than in basal (t-test)


```{r ttestOnStat, echo=FALSE,cache=TRUE}
apply(DEmat$ramp$stat,1,function(statRow){
  res <- t.test(x = statRow[1:2],y = statRow[3:5],alternative = "greater")
  return(c(pVal=res$p.value, tStat=res$statistic))
}) -> rampCoreVsBasalttest

apply(DEmat$peak$stat,1,function(statRow){
  res <- t.test(x = statRow[1:2],y = statRow[3:5],alternative = "greater")
  return(c(pVal=res$p.value, tStat=res$statistic))
}) -> peakCoreVsBasalttest

```


```{r tstatCoreVsBasal, echo=FALSE}
idx <- order(peakCoreVsBasalttest["tStat.t",],decreasing = T)[1:40]

heatmap(DEmat$peak$stat[idx,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13),
        RowSideColors = ifelse(pamlMat[idx,"H4b"]<0.05,"red","grey"),
        scale="none", Colv = NA,main="Peak up in core (H4b)")

tmpTbl <- cbind(grpAnnotTbl[rownames(DEmat$peak$stat)[idx], ],
                tStat=round(sort(peakCoreVsBasalttest["tStat.t",],decreasing = T)[1:40],digits = 3))
DT::datatable(tmpTbl)


idx <- order(rampCoreVsBasalttest["tStat.t",],decreasing = T)[1:40]

heatmap(DEmat$ramp$stat[idx,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13),
        RowSideColors = ifelse(pamlMat[idx,"H4b"]<0.05,"red","grey"),
        scale="none", Colv = NA,main="Ramp up in core (H4b)")

tmpTbl <- cbind(grpAnnotTbl[rownames(DEmat$ramp$stat)[idx], ],
                tStat=round(sort(rampCoreVsBasalttest["tStat.t",],decreasing = T)[1:40],digits = 3))
DT::datatable(tmpTbl)
```

### Up in all except Stipa (StLa)

```{r upInAllExceptStLa, echo=FALSE}
notStLa <- c("BrDi", "HoVu", "MeNu1", "NaSt")
whichNotStLaPeakUp <- which(apply(sigMat$peak[ ,notStLa],1,all) & !sigMat$peak[ ,"StLa"])

heatmap(DEmat$peak$stat[whichNotStLaPeakUp,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13),
        scale="none", Rowv = NA, Colv = NA, main="Peak up in all except Stipa")

DT::datatable(grpAnnotTbl[names(whichNotStLaPeakUp), ])


whichNotStLaRampUp <- which(apply(sigMat$ramp[ ,notStLa],1,all) & !sigMat$ramp[ ,"StLa"])

heatmap(DEmat$ramp$stat[whichNotStLaRampUp,], breaks=c(-30,seq(-5.5,5.5,by=1),30),
        col=colorRampPalette(c("green","black","red"))(13), Rowv = NA,
        scale="none", Colv = NA, main="Ramp up in all except Stipa")

DT::datatable(grpAnnotTbl[names(whichNotStLaRampUp), ])
```

> NOTE: StLa does not have a T-1 sample and the ramp effect is calculated as the contrast between T0 (instead of T-1) and T3/4. Since T0 and T3/4 was sampled at a different time of the day it could be confounded with effects of the daily rythm.

## Lineage specific

### Nardus specific response

```{r upInNaSt, echo=FALSE}

whichNaStPeakUp <- which( rowSums(sigMat$peak)==1 & sigMat$peak[ ,"NaSt"])

heatmap(DEmat$peak$stat[whichNaStPeakUp,], breaks=c(-30,seq(-5.5,5.5,by=1),30),
        col=colorRampPalette(c("green","black","red"))(13), #Rowv = NA,
        RowSideColors = ifelse(pamlMat[whichNaStPeakUp,"H5a"]<0.05,"red","grey"),        
        scale="none", Colv = NA, main="Peak up in Nardus only (H5a)")

summary(table(H5a = pamlMat[ ,"H5a"]<0.05, 
              NaStPeakUp = allGrpIDs %in% names(whichNaStPeakUp)))

DT::datatable(grpAnnotTbl[names(whichNaStPeakUp), ])

DT::datatable(testGOgrps(names(whichNaStPeakUp),bkGrpIDs = rownames(sigMat$peak)))


whichNaStRampUp <- which( rowSums(sigMat$ramp)==1 & sigMat$ramp[ ,"NaSt"])

heatmap(DEmat$ramp$stat[whichNaStRampUp,], breaks=c(-30,seq(-5.5,5.5,by=1),30),
        col=colorRampPalette(c("green","black","red"))(13), #Rowv = NA,
        RowSideColors = ifelse(pamlMat[whichNaStRampUp,"H5a"]<0.05,"red","grey"),        
        scale="none", Colv = NA, main="Ramp up in Nardus only (H5a)")

summary(table(H5a = pamlMat[ ,"H5a"]<0.05, 
              NaStPeakUp = allGrpIDs %in% names(whichNaStRampUp)))


DT::datatable(grpAnnotTbl[names(whichNaStRampUp), ])

DT::datatable(testGOgrps(names(whichNaStRampUp),bkGrpIDs = rownames(sigMat$peak)))
```

### Melica specific response

```{r upInMeNu, echo=FALSE}

whichMeNuPeakUp <- which( rowSums(sigMat$peak)==1 & sigMat$peak[ ,"MeNu1"])

heatmap(DEmat$peak$stat[whichMeNuPeakUp,], breaks=c(-30,seq(-5.5,5.5,by=1),30),
        col=colorRampPalette(c("green","black","red"))(13), #Rowv = NA,
        RowSideColors = ifelse(pamlMat[whichMeNuPeakUp,"H5b"]<0.05,"red","grey"),        
        scale="none", Colv = NA, main="Peak up in Melica only (H5b)")

summary(table(H5a = pamlMat[ ,"H5b"]<0.05, 
              NaStPeakUp = allGrpIDs %in% names(whichMeNuPeakUp)))

DT::datatable(grpAnnotTbl[names(whichMeNuPeakUp), ])

DT::datatable(testGOgrps(names(whichMeNuPeakUp),bkGrpIDs = rownames(sigMat$peak)))

whichMeNuRampUp <- which( rowSums(sigMat$ramp)==1 & sigMat$ramp[ ,"MeNu1"])

heatmap(DEmat$ramp$stat[whichMeNuRampUp,], breaks=c(-30,seq(-5.5,5.5,by=1),30),
        col=colorRampPalette(c("green","black","red"))(13), #Rowv = NA,
        RowSideColors = ifelse(pamlMat[whichMeNuRampUp,"H5b"]<0.05,"red","grey"),        
        scale="none", Colv = NA, main="Ramp up in Melica only")

summary(table(H5a = pamlMat[ ,"H5b"]<0.05, 
              NaStPeakUp = allGrpIDs %in% names(whichMeNuRampUp)))


DT::datatable(grpAnnotTbl[names(whichMeNuRampUp), ])

DT::datatable(testGOgrps(names(whichMeNuRampUp),bkGrpIDs = rownames(sigMat$peak)))

```

### Test for GO enrichment of groups with positive selection

#### H4a - positive selection in root of pooid branch

```{r pamlGO_H4a, echo=FALSE}
isNA <- is.na(pamlMat[ ,"H5a"])
DT::datatable( testGOgrps( allGrpIDs[pamlMat[ !(isNA),"H5a"]<0.05],
                           bkGrpIDs = allGrpIDs[!isNA]) )
```

#### H4b - positive selection in root of core pooid branch (Lolium + Hordeum + Brachypodium)

```{r pamlGO_H4b, echo=FALSE}
DT::datatable( testGOgrps( allGrpIDs[pamlMat[ ,"H4b"]<0.05], bkGrpIDs = allGrpIDs) )
```

#### H4c - positive selection in root of inner core pooid branch (Lolium + Hordeum)
```{r pamlGO_H4c, echo=FALSE}
DT::datatable( testGOgrps( allGrpIDs[pamlMat[ ,"H4c"]<0.05], bkGrpIDs = allGrpIDs) )
```

#### H5a - positive selection in Nardus
```{r pamlGO_H5a, echo=FALSE}
DT::datatable( testGOgrps( allGrpIDs[pamlMat[ ,"H5a"]<0.05], bkGrpIDs = allGrpIDs) )
```

#### H5b - positive selection in Melica
```{r pamlGO_H5b, echo=FALSE}
DT::datatable( testGOgrps( allGrpIDs[pamlMat[ ,"H5b"]<0.05], bkGrpIDs = allGrpIDs) )
```


```{r echo=FALSE, eval=FALSE}
SE <- DEnew$HoVu$resPeak$lfcSE
x <- DEnew$HoVu$resPeak$log2FoldChange
z <- DEnew$HoVu$resPeak$stat
p <- DEnew$HoVu$resPeak$padj

# z-score of 2 is equivalent to P-adjusted of 0.05

table(over0 = z>2, under1 = ((1-x)/SE) > 2)
hist(x[(z>2) & !(((1-x)/SE) > 2)])

hist(SE)
plot(x,SE, col=1+1*(x>2*SE)+2*((1-x)>2*SE))

table(1*(x>2*SE)+2*((1-x)>2*SE))

spc2col <- c(BrDi="cyan",HoVu="blue",MeNu1="red",StLa="yellow",NaSt="black")
plot(NULL,xlim=c(0,1.5),ylim=c(0,4))
for(spc in names(DEnew)){
  lines(density(DEnew[[spc]]$resRamp$lfcSE),col=spc2col[spc])  
}
cor(DEnew$BrDi$resRamp$lfcSE,DEnew$StLa$resRamp$baseMean)
plot(density(log2(DEnew$StLa$resRamp$baseMean[ DEnew$StLa$resRamp$lfcSE > 0.9])))
lines(density(log2(DEnew$StLa$resRamp$baseMean[ DEnew$StLa$resRamp$lfcSE < 0.9])),col="red")


```

