---
title: "GO analysis using Brachypodium Annotations"
output:
  html_document:
    toc: yes
---

```{r loadLibraries, echo=F, message=F}
library(DESeq2)
library(gplots)
# # un-comment to install from bioconductor:
# source("http://bioconductor.org/biocLite.R")
# biocLite("topGO")
library(topGO)


source("~/GRewd/pipeline/R/orthoGrpTools.R")
```


```{r loadGOannot, echo=FALSE}
GOannotFile <- "~/GRewd/pipeline/indata/Bd_GO.mart_export.txt"

# read GO annotations
BdGOtbl <- read.table(GOannotFile,sep="\t", quote = "", stringsAsFactors=F,header = T)

# convert to gene2GO list
BdGene2GO <- split(BdGOtbl$GO.term.accession,BdGOtbl$Gene.stable.ID)
```

```{r getPlantSlimGOs, echo=FALSE}
plantSlim.txt <- readLines("http://www.geneontology.org/ontology/subsets/goslim_plant.obo")
plantSlimGOs <- sub("^id: ","",plantSlim.txt[grep("^\\[Term\\]$",plantSlim.txt)+1])
```


```{r loadData, echo=FALSE}
orthoPath <- "/mnt/NOBACKUP/mariansc/share/orthos"


# load DEmat based based on superGenes
DEmat <- readRDS("~/GRewd/pipeline/data/DEmat.RDS")  # from: superGeneModel.Rmd

# load groups
goodGrps <- loadOrthoGrpsArray(file.path(orthoPath,"splitGroups/goodGroups.txt"))

# get names of groups in supergene table
superGrpIDs <- rownames(DEmat$peak$pVal)
```


```{r defGOFunctions, echo=FALSE}

BRADIgeneIDs2grpIDs <- function(geneIDs, grps){
  flatGrpIDs <- rep(rownames(grps),sapply(grps[ ,"Bd_R"],length))
  spcGeneIDs <- unlist(grps[,"Bd_R"],use.names = F)

  BRADIgeneIDs <- sub("\\.[0-9]+$","",toupper(spcGeneIDs))
  flatGrpIDs[match(geneIDs,BRADIgeneIDs)]
}

# make grp2GO list
BdGOtbl$grpID <- BRADIgeneIDs2grpIDs(BdGOtbl$Gene.stable.ID,goodGrps)

grpID2GO <- lapply( split( BdGOtbl$GO.term.accession, BdGOtbl$grpID ), unique )


getGOterms <- function(GOids){
  
  quotedGOids <- paste(paste("'", unique(GOids), "'", sep = ""), collapse = ",")
  retVal <- dbGetQuery(GO_dbconn(), 
                       paste0("SELECT term, go_id FROM go_term ",
                              "WHERE go_id IN (", quotedGOids, ");"))
  return( setNames(retVal$term,retVal$go_id)[GOids] )
}

testGOgrps <- function(grpIDs, bkGrpIDs = NULL, ontology = "BP"){
  # get annotated background groups
  if( !is.null(bkGrpIDs) ){
    annBkGrpIDs <- intersect(names(grpID2GO),bkGrpIDs)
  } else {
    annBkGrpIDs <- names(grpID2GO)
  }
  
  allGenes <- factor(as.integer(annBkGrpIDs %in% grpIDs))
  names(allGenes) <- annBkGrpIDs
  
  sink("/dev/null") # silence the output from topGO
  
  GOdata <- new("topGOdata", ontology = ontology, allGenes = allGenes,
                annot = annFUN.gene2GO, gene2GO = grpID2GO)
  
  
  # run fisher test using classic algorithm
  classic.fisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")
  weight.fisher <- runTest(GOdata, algorithm = "weight", statistic = "fisher")
  
  sink() # stop sinking output
  
#   GenTable(GOdata, classic = classic.fisher, weight=weight.fisher,
#            orderBy = "classic", ranksOf="classic", topNodes = 10)
  tbl <- termStat(GOdata)
  tbl$pval.classic.fisher <- score(classic.fisher)[rownames(tbl)]
  tbl$pval.weight.fisher <- score(weight.fisher)[rownames(tbl)]


  # get grps belonging to each of the 
  GO2grpID <- genesInTerm(GOdata)
  g_GO2grpID <<- GO2grpID # store in global variable for later

  tbl$SigGrpIDs <- ""
  idx <- tbl$Significant > 0 & tbl$Significant < 50
  tbl$SigGrpIDs[idx] <- sapply(GO2grpID[ rownames(tbl)[idx] ], function(GOgrpID){
    paste(GOgrpID[GOgrpID %in% grpIDs], collapse=",")
  })

  return(tbl)
}
```

### DE supergenes

Based on the "super-genes" expression values, the genes are called DE if either peak or ramp are DE (p.adj<0.05 & |lfc|>1)

```{r sigVennPlet, echo=FALSE}
# Get significant genes for peak and ramp (NA values equals not significant)
sigMatUpDown <- lapply(DEmat, with, { ifelse(is.na(pAdj), F, pAdj < 0.05 & abs(lfc) > 1) })
# Combine peak and ramp (significant if significant in either peak or ramp)
combSigMat <- sigMatUpDown$peak | sigMatUpDown$ramp

# plot Venn diagram
venn(apply(combSigMat,2,which))
title(main="significant peak or ramp (up or down)",sub="p.adj<0.05 & |lfc|>1")
```

```{r calcSig, echo=FALSE}
# get species specific LTI genes
spcs <- colnames(combSigMat)
sapply( spcs, function(spc){
  # significant in spc and not significant in any other spc 
  combSigMat[ ,spc] & !apply(combSigMat[ ,spcs != spc],1,any)
}) -> spcSig


# Species included in each branch/hypothesis (excluding lolium)
hSpcs2 <- list(
  H4b = c("HoVu","BrDi"),
  H4e = c("HoVu","BrDi","MeNu1"),
  H4d = c("HoVu","BrDi","MeNu1","StLa"),
  H4a = c("HoVu","BrDi","MeNu1","StLa","NaSt")
)

# Get branch specific LTI genes
sapply( hSpcs2, function(branchSpcs){
  # significant in all branch species and not significant in any other species 
  apply(combSigMat[ ,spcs %in% branchSpcs],1,all) & 
    !apply(combSigMat[ ,!(spcs %in% branchSpcs),drop=F],1,any)
}) -> hSig

```




```{r DE_AllGO, echo=F, cache=TRUE}

# run topGO on all branch and species specific DE genes

apply(cbind(hSig,spcSig),2,function(sig){
# apply(hSig,2,function(sig){
  testGOgrps(grpIDs = superGrpIDs[sig], bkGrpIDs = superGrpIDs)
}) -> sigGO

```

```{r plotGObarplot, echo=FALSE, fig.height=9,fig.width=12, eval=FALSE}

# get all significantly enriched GO's (weight.fisher P<0.05)
allSigGOs <- unique(unlist(lapply(sigGO,function(tbl){
#   rownames(tbl)[which(tbl$pval.weight.fisher<0.05)]  
  rownames(tbl)[which(tbl$pval.classic.fisher<0.05)]  
})))

# make matrix number of DE genes per GO
sigGOmat <- cbind(sapply(sigGO,with,Significant),all=sigGO$H4a$Annotated)
rownames(sigGOmat) <- rownames(sigGO$H4a)

# sigGOpvalmat <- cbind(sapply(sigGO,with,pval.weight.fisher))
sigGOpvalmat <- cbind(sapply(sigGO,with,pval.classic.fisher))
rownames(sigGOpvalmat) <- rownames(sigGO$H4a)



# GO's sorted by number of occurences in background
sortedSigGOs <- names(sort(sigGOmat[allSigGOs,"all"],decreasing = T))
sortedSigGOs <- sortedSigGOs[sortedSigGOs %in% plantSlimGOs]

# Give a color to each of the significant GO's
GO2Col <- setNames(RColorBrewer::brewer.pal(n = min(length(sortedSigGOs),12),name = "Set3"),
                   sortedSigGOs[1:min(length(sortedSigGOs),12)])
# GO2Col <- setNames(RColorBrewer::brewer.pal(n = 12,name = "Set3"),
#                    sortedSigGOs[1:12])

sigGOnormMat <- t(apply(sigGOmat[sortedSigGOs,],1,"/",colSums(sigGOmat)))

barplot(sigGOnormMat,col=GO2Col[sortedSigGOs],width=0.5, space=1,
        las=2, xlim=c(0,30), ylab="Proportion of GO annotations")
# legend("right",
#        legend = c("(div GO's)",
#                   rev(getGOterms(sortedSigGOs[1:min(length(sortedSigGOs),12)]))),
#        fill=c("white",rev(GO2Col)))
legend("right",
       legend = rev(getGOterms(sortedSigGOs[1:min(length(sortedSigGOs),12)])),
       fill=rev(GO2Col))

y <- apply(sigGOnormMat,2,function(x){cumsum(x)-x/2})[,1:9]
pval <- sigGOpvalmat[sortedSigGOs, ]
sigLabel <- cut(pval,c(0,0.005,0.01,0.05,1),labels=c("***","**","*",""))

text(x = rep(1:9,each=nrow(y)),y=y[,1:9],labels = sigLabel,adj = 0)

# legend( "bottomleft",#inset = c(0, -0.2)
#         bty = "n", 
#         x.intersp=0.5, xjust=0,yjust=0, 
#         legend=c("***   P<0.005", " **   P<0.01", "  *   P<0.05"),
#         cex = 0.75, xpd = TRUE, horiz=T)
```



```{r GOtable1, echo=FALSE, eval=FALSE}

# make a table containing all GO tests
# put hypothesis in first column and GO in second

Reduce(rbind,lapply(names(sigGO), function(h){
  tmp <- sigGO[[h]][ sigGO[[h]]$pval.weight.fisher<0.05, ]
  
  cbind(GO=rownames(tmp),hyp=rep(h,nrow(tmp)),tmp)
})) -> combinedGOtbls

combinedGOtbls$pval <- signif(x = combinedGOtbls$pval.weight.fisher,2)
combinedGOtbls$pval.classic.fisher <- NULL
combinedGOtbls$pval.weight.fisher <- NULL
combinedGOtbls$term <- getGOterms(as.character(combinedGOtbls$GO))

combinedGOtbls[1:5,]

DT::datatable(combinedGOtbls,rownames = F)
```




### Up or down

We want to look specifically at genes that are up regulated or down regulated in response to cold. The genes are called DE up (or down) if either peak or ramp are DE up (or down) (p.adj<0.05 & (-)lfc>1).

```{r sigVennPlotUpDown, echo=FALSE}
# Get significant genes for peak and ramp (NA values equals not significant)

sigDE <- lapply(DEmat, with, { ifelse(is.na(pAdj), F, pAdj < 0.05 & lfc > 1) })
sigMatUp <- sigDE$peak | sigDE$ramp
sigDE <- lapply(DEmat, with, { ifelse(is.na(pAdj), F, pAdj < 0.05 & lfc < -1) })
sigMatDown <- sigDE$peak | sigDE$ramp
rm(sigDE)

# plot Venn diagram
venn(apply(sigMatUp,2,which))
title(main="significant peak or ramp UP",sub="p.adj<0.05 & lfc > 1")

venn(apply(sigMatDown,2,which))
title(main="significant peak or ramp DOWN",sub="p.adj<0.05 & lfc < -1")

```



```{r defFun2, echo=FALSE}

runGOallHyp <- function(sigMat){
  spcs <- colnames(sigMat)
  sapply( spcs, function(spc){
    # significant in spc and not significant in any other spc 
    sigMat[ ,spc] & !apply(sigMat[ ,spcs != spc],1,any)
  }) -> spcSig
  
  
  # Species included in each branch/hypothesis (excluding lolium)
  hSpcs2 <- list(
    H4b = c("HoVu","BrDi"),
    H4e = c("HoVu","BrDi","MeNu1"),
    H4d = c("HoVu","BrDi","MeNu1","StLa"),
    H4a = c("HoVu","BrDi","MeNu1","StLa","NaSt")
  )
  
  # Get branch specific LTI genes
  sapply( hSpcs2, function(branchSpcs){
    # significant in all branch species and not significant in any other species 
    apply(sigMat[ ,spcs %in% branchSpcs],1,all) & 
      !apply(sigMat[ ,!(spcs %in% branchSpcs),drop=F],1,any)
  }) -> hSig
  
  
  # run topGO on all branch and species specific DE genes
  
  apply(cbind(hSig,spcSig),2,function(sig){
#   apply(hSig,2,function(sig){
    testGOgrps(grpIDs = superGrpIDs[sig], bkGrpIDs = superGrpIDs)
  })
}
```

```{r calcSigUpDown,echo=FALSE, cache=TRUE}
sigGOup <- runGOallHyp(sigMat = sigMatUp)
sigGOdown <- runGOallHyp(sigMat = sigMatDown)

```

```{r barPlotGOUpDown, echo=FALSE, fig.height=9,fig.width=12, eval=FALSE}

sigGOupDown <- c( setNames(sigGOup,paste0(names(sigGOup),"Up")),
            setNames(sigGOdown,paste0(names(sigGOdown),"Down")))


# get all significantly enriched GO's (weight.fisher P<0.05)
allSigGOs <- unique(unlist(lapply(sigGOupDown,function(tbl){
#   rownames(tbl)[which(tbl$pval.weight.fisher<0.05)]  
  rownames(tbl)[which(tbl$pval.classic.fisher<0.05)]  
})))


# make matrix number of DE genes per GO
sigGOmat <- cbind(sapply(sigGOupDown,with,Significant),all=sigGOupDown[[1]]$Annotated)
rownames(sigGOmat) <- rownames(sigGOupDown[[1]])

# sigGOpvalmat <- cbind(sapply(sigGOupDown,with,pval.weight.fisher))
sigGOpvalmat <- cbind(sapply(sigGOupDown,with,pval.classic.fisher))
rownames(sigGOpvalmat) <- rownames(sigGOupDown[[1]])

# GO's sorted by number of occurences in background
sortedSigGOs <- names(sort(sigGOmat[allSigGOs,"all"],decreasing = T))
sortedSigGOs <- sortedSigGOs[sortedSigGOs %in% plantSlimGOs]

sortedSigGOs <- sortedSigGOs[3:length(sortedSigGOs)]

# Give a color to each of the significant GO's
GO2Col <- setNames(c(RColorBrewer::brewer.pal(n = 12,name = "Paired"),"white"),
                   sortedSigGOs[1:13])

sigGOnormMat <- t(apply(sigGOmat[sortedSigGOs,],1,"/",colSums(sigGOmat)))


sigGOnormMat <- sigGOnormMat[,]


barplot(sigGOnormMat,col=GO2Col[sortedSigGOs],width=0.5, space=1,
        las=2, xlim=c(0,30), ylab="Proportion of GO annotations")
legend("right",legend = rev(getGOterms(sortedSigGOs)),
       fill=rev(GO2Col))
# legend("right",legend = c("(div. GO's)",rev(getGOterms(sortedSigGOs[1:12]))),
#        fill=c("white",rev(GO2Col)))

y <- apply(sigGOnormMat,2,function(x){cumsum(x)-x/2})[,-ncol(sigGOnormMat)]
pval <- sigGOpvalmat[sortedSigGOs, ]
sigLabel <- cut(pval,c(0,0.005,0.01,0.05,1),labels=c("***","**","*",""))

text(x = rep(1:ncol(y),each=nrow(y)),y=y[,1:ncol(y)],labels = sigLabel,adj = 0)

# legend( "topleft",
#         inset = c(0, -0.2), bty = "n", 
#         x.intersp=0.5, xjust=0,yjust=0, 
#         legend=c("***   P<0.005", " **   P<0.01", "  *   P<0.05"),
#         cex = 0.75, xpd = TRUE, horiz=T)

```

### GO enrichment for species specific and clade specific DE genes


```{r comboBarplot, echo=FALSE, fig.height=9,fig.width=14}

# get all significantly enriched GO's (fisher P<0.05)

sigGOcomb <- c( sigGO,
            setNames(sigGOup,paste0(names(sigGOup),"Up")),
            setNames(sigGOdown,paste0(names(sigGOdown),"Down")))

# remove H4e hypothesis
sigGOcomb <- sigGOcomb[!grepl("^H4e",names(sigGOcomb))]

combSigGOs <- unique(unlist(lapply(sigGOcomb,function(tbl){
  rownames(tbl)[which(tbl$pval.classic.fisher<0.05)]  
})))

# make matrix number of DE genes per GO
sigGOmat <- cbind(sapply(sigGOcomb,with,Significant),all=sigGOcomb[[1]]$Annotated)
rownames(sigGOmat) <- rownames(sigGOcomb[[1]])

sigGOpvalmat <- cbind(sapply(sigGOcomb,with,pval.classic.fisher))
rownames(sigGOpvalmat) <- rownames(sigGOcomb[[1]])

# GO's sorted by number of occurences in background
sortedSigGOs <- names(sort(sigGOmat[combSigGOs,"all"],decreasing = T))
# only keep GOs from plantSlim
sortedSigGOs <- sortedSigGOs[sortedSigGOs %in% plantSlimGOs]

# remove some specific GOs that are redundant
redundantGOs <- c("GO:0008152","GO:0009058","GO:0007154","GO:0009875")
sortedSigGOs <- sortedSigGOs[!(sortedSigGOs %in% redundantGOs)]

# Give a color to each of the significant GO's
GO2Col <- setNames(c(RColorBrewer::brewer.pal(n = 12,name = "Paired"),"white","black",
                     RColorBrewer::brewer.pal(n = 8,name = "Dark2")),
                   sortedSigGOs)

idxSlim <- rownames(sigGOmat) %in% plantSlimGOs[!(plantSlimGOs %in% redundantGOs)]


sigGOnormMat <- t(apply(sigGOmat[sortedSigGOs,],1,"/",colSums(sigGOmat[idxSlim, ])))


# reorder columns
colIdx <- c(as.vector(t(matrix(1:(ncol(sigGOnormMat)-1),ncol=3))),ncol(sigGOnormMat))
sigGOnormMat <- sigGOnormMat[,colIdx]

idxSigGO <- rownames(sigGOmat)%in%sortedSigGOs
cbind(
  # total number of annotations
  nAnn = colSums(sigGOmat[idxSlim, ]),
  # total number of significant annotations
  nSigAnn = colSums(sigGOmat[idxSigGO, ]),
  # count number of genes with annotations
  gAnn = c(sapply(sigGOcomb, with, length(unique(unlist(strsplit(SigGrpIDs[idxSlim],","))))),NA),
  # count number of genes with significant annotations
  gSigAnn = c(sapply(sigGOcomb, with, length(unique(unlist(strsplit(SigGrpIDs[idxSigGO],","))))),
              NA) # length(unique(unlist(GO2grpID[sortedSigGOs])))
)[colIdx,] -> colStats

labs <- paste0(rownames(colStats),"(",colStats[,"nSigAnn"],"/",colStats[,"gSigAnn"],")")

par(mar=c(8, 4, 4, 2) + 0.1)
barplot(sigGOnormMat,col=GO2Col[sortedSigGOs],width=0.5, space=c(1,rep(c(1,1,2),8)),
        las=2, xlim=c(0,50), ylab="Proportion of slim GO annotations",
        names.arg=labs)
par(mar=c(5, 4, 4, 2) + 0.1)
legend("topright",legend = rev(getGOterms(sortedSigGOs)),
       fill=rev(GO2Col[sortedSigGOs]))
# legend("right",legend = c("(div. GO's)",rev(getGOterms(sortedSigGOs[1:12]))),
#        fill=c("white",rev(GO2Col)))

y <- apply(sigGOnormMat,2,function(x){cumsum(x)-x/2})[,-ncol(sigGOnormMat)]
pval <- sigGOpvalmat[rownames(sigGOnormMat),colIdx[-25]]
sigLabel <- cut(pval,c(0,0.005,0.01,0.05,1),labels=c("***","**","*",""))

text(x = 0.2+rep(cumsum(c(1,rep(c(1,1,1.5),8)))[-25],each=nrow(y)),
     y=y[,1:ncol(y)],labels = sigLabel,adj = 0.5,srt=90)

```

```{r comboBarplotNoSpcs, echo=FALSE, fig.height=9,fig.width=14}

# get all significantly enriched GO's (fisher P<0.05)

sigGOcomb <- c( sigGO,
            setNames(sigGOup,paste0(names(sigGOup),"Up")),
            setNames(sigGOdown,paste0(names(sigGOdown),"Down")))

# remove H4e hypothesis
sigGOcomb <- sigGOcomb[!grepl("^H4e",names(sigGOcomb))]

# Keep only branch hypotheses
sigGOcomb <- sigGOcomb[grepl("^H4",names(sigGOcomb))]


combSigGOs <- unique(unlist(lapply(sigGOcomb,function(tbl){
  rownames(tbl)[which(tbl$pval.classic.fisher<0.05)]  
})))

# make matrix with number of DE genes per GO
sigGOmat <- cbind(sapply(sigGOcomb,with,Significant),all=sigGOcomb[[1]]$Annotated)
rownames(sigGOmat) <- rownames(sigGOcomb[[1]])

sigGOpvalmat <- cbind(sapply(sigGOcomb,with,pval.classic.fisher))
rownames(sigGOpvalmat) <- rownames(sigGOcomb[[1]])

# GO's sorted by number of occurences in background
sortedSigGOs <- names(sort(sigGOmat[combSigGOs,"all"],decreasing = T))
# only keep GOs from plantSlim
sortedSigGOs <- sortedSigGOs[sortedSigGOs %in% plantSlimGOs]

# remove some specific GOs that are redundant
redundantGOs <- c("GO:0008152","GO:0009058","GO:0007154","GO:0009875")
sortedSigGOs <- sortedSigGOs[!(sortedSigGOs %in% redundantGOs)]

# Give a color to each of the significant GO's
GO2Col <- setNames(c(RColorBrewer::brewer.pal(n = 12,name = "Paired"),"white","black",
                     RColorBrewer::brewer.pal(n = 8,name = "Dark2")),
                   sortedSigGOs)

idxSlim <- rownames(sigGOmat) %in% plantSlimGOs[!(plantSlimGOs %in% redundantGOs)]


sigGOnormMat <- t(apply(sigGOmat[sortedSigGOs,],1,"/",colSums(sigGOmat[idxSlim, ])))


# reorder columns
colNamesOrdered <- c("H4a","H4aUp","H4aDown","H4d","H4dUp","H4dDown","H4b","H4bUp","H4bDown","all")
colDisplayNamesOrdered <- c(rep(c("Up/Down","Up only","Down only"),3),"all")

colOrder <- match(colNamesOrdered,colnames(sigGOnormMat))

idxSigGO <- rownames(sigGOmat)%in%sortedSigGOs
cbind(
  # total number of annotations
  nAnn = colSums(sigGOmat[idxSlim, ]),
  # total number of significant annotations
  nSigAnn = colSums(sigGOmat[idxSigGO, ]),
  # count number of genes with annotations
  gAnn = c(sapply(sigGOcomb, with, length(unique(unlist(strsplit(SigGrpIDs[idxSlim],","))))),NA),
  # count number of genes with significant annotations
  gSigAnn = c(sapply(sigGOcomb, with, length(unique(unlist(strsplit(SigGrpIDs[idxSigGO],","))))),
              # calculate number of genes that are annotated with at least one of the sortedSigGOs:
              length(unique(unlist(g_GO2grpID[sortedSigGOs]))))
) -> colStats


labsOrdered <- paste0(colDisplayNamesOrdered,
                      " (",colStats[colOrder,"nSigAnn"],"/",colStats[colOrder,"gSigAnn"],")")


svg(filename = "~/GRewd/pipeline/articleFigures/FigureGO.svg",
    width = 9, height = 6)

par(mar=c(8, 4, 4, 2) + 0.1)
barplot(sigGOnormMat[,colOrder],col=GO2Col[sortedSigGOs],width=0.5, space=c(1,rep(c(1,1,2),3)),
        las=2, xlim=c(0,30), ylab="Proportion of GO annotations",
        names.arg=labsOrdered)
par(mar=c(5, 4, 4, 2) + 0.1)
legend("topright",legend = rev(getGOterms(sortedSigGOs)),
       fill=rev(GO2Col[sortedSigGOs]))
# legend("right",legend = c("(div. GO's)",rev(getGOterms(sortedSigGOs[1:12]))),
#        fill=c("white",rev(GO2Col)))

y <- apply(sigGOnormMat[,colOrder],2,function(x){cumsum(x)-x/2})[,-ncol(sigGOnormMat)]
pval <- sigGOpvalmat[rownames(sigGOnormMat),colOrder[-ncol(sigGOnormMat)]]
sigLabel <- cut(pval,c(0,0.005,0.01,0.05,1),labels=c("***","**","*",""))

text(x = 0.2+rep(cumsum(c(1,rep(c(1,1,1.5),3)))[-ncol(sigGOnormMat)],each=nrow(y)),cex = 0.8,
     y=y[,1:ncol(y)],labels = sigLabel,adj = 0.5,srt=90)

invisible(dev.off())

```


GO annotations were slimmed using plantSlim to reduce the number annotations to more general terms. In adddition, four more GOs were removed to reduce redundancy. Only the biological process GOs were considered. The barplot shows the GOs that were significantly enriched in at least one of the DE gene sets (Fisher's exact test, P<0.05). The height of the bars reflects the number of annotations in each DE gene set as the proportion of the total number of annotations in the same gene set. As a reference the rightmost bar shows the proportion of GO annotations for all genes.

In paranthesis are the number of annotations and corresponding number of genes (Note that there can be several annotations per gene) that are among the GOs shown in the barplot.

The number of stars next to a bar in the plot reflects the P-value of the corresponding GO enrichment test (1: P<0.05, 2: P<0.01, 3: P<0.005).


```{r deadCode, echo=FALSE, eval=FALSE}
getAllParentGOs <- function(GOids){
  df <- toTable(GOBPPARENTS)
  lapply(setNames(GOids,GOids), function(GOid){
    GOparentIds <- df[df[,1] %in% GOid,2]
    getGOterms(GOparentIds)
    allParentIds  <- GOparentIds
    while( length(GOparentIds) > 0){
      length(GOparentIds) > 0
      GOparentIds <- unique(df[df[,1] %in% GOparentIds,2])
      allParentIds  <- c(allParentIds,GOparentIds)
    }
    unique(allParentIds)
  })
}
allParentGOs$`GO:0007154`
allParentGOs <- getAllParentGOs(GOids = sortedSigGOs)
sapply(sortedSigGOs, function(GOid){
  which(sapply(allParentGOs, function(parentGOs){GOid %in% parentGOs}))
})
getGOterms(c("GO:0007154","GO:0007165", "GO:0009875"))
```

### Table of all significant GO 
Use the search box to limit view to single hypothesis or term.

Click the the number of significant genes to view details of those genes.

```{r GOtableUpDown, echo=FALSE}

# make a table containing all GO tests
# put hypothesis in first column and GO in second

sigGOall <- c( setNames(sigGO,paste0(names(sigGO),"_")),
            setNames(sigGOup,paste0(names(sigGOup),"Up")),
            setNames(sigGOdown,paste0(names(sigGOdown),"Down")))


Reduce(rbind,lapply(names(sigGOall), function(h){
#   tmp <- sigGOall[[h]][ sigGOall[[h]]$pval.weight.fisher<0.05, ]
  tmp <- sigGOall[[h]][ sigGOall[[h]]$pval.classic.fisher<0.05 | 
                        sigGOall[[h]]$pval.weight.fisher<0.05, ]
  cbind(GO=rownames(tmp),hyp=rep(h,nrow(tmp)),tmp)
})) -> combinedGOtbls

combinedGOtbls$Pval <- signif(x = combinedGOtbls$pval.classic.fisher,2)
combinedGOtbls$`Pval(weight)` <- signif(x = combinedGOtbls$pval.weight.fisher,2)
combinedGOtbls$pval.classic.fisher <- NULL
combinedGOtbls$pval.weight.fisher <- NULL
combinedGOtbls$slim <- ifelse(as.character(combinedGOtbls$GO) %in% plantSlimGOs,"X","")

combinedGOtbls$term <- getGOterms(as.character(combinedGOtbls$GO))

combinedGOtbls$Significant <- ifelse(combinedGOtbls$SigGrpIDs=="", 
                                     as.character(combinedGOtbls$Significant),
                                     paste(
'<a href="http://rstudio-01.orion.nmbu.no:3838/lagr/SuperGeneList/?grpIDs=', 
                           combinedGOtbls$SigGrpIDs, '">',combinedGOtbls$Significant,'</a>', sep=""))
combinedGOtbls$SigGrpIDs <- NULL

DT::datatable(combinedGOtbls,rownames = F,escape=F)
```

