---
title: "Paml results overview"
output: html_document
---


```{r loadLibraries, echo=FALSE, message=FALSE}
library(DESeq2)
library(gplots)
library(stringr)


source("~/GRewd/pipeline/R/orthoGrpTools.R")
```

```{r loadData, echo=FALSE}
orthoPath <- "/mnt/NOBACKUP/mariansc/share/orthos"


pamlResFile<-"~/GRewd/pipeline/data/pamlRes.RDS"
DEnewFile<-"~/GRewd/pipeline/data/DEnew.RDS"
VSTnewFile<-"~/GRewd/pipeline/data/VSTnew.RDS"
superExprMatrixFile <- "~/GRewd/pipeline/data/superExprMatrix.RDS"

# load DE
load(file.path(orthoPath,"DESeq/DE.RData"))

# load superGenes expression
expr <- readRDS(file = superExprMatrixFile)  
# load DE based and vst based on superGenes
DEnew <- readRDS(DEnewFile)  
vst <- readRDS(VSTnewFile)  


# convert DE results to matrix format
lapply( list( peak="resPeak", ramp="resRamp"), function( peakOrRamp ){
  lapply( list( pVal="pvalue", pAdj="padj", stat="stat",lfc="log2FoldChange"), function( stat ){
    m <- sapply(DEnew, function(DEspc){DEspc[[peakOrRamp]][[stat]]})
    rownames(m) <- rownames(expr)
    return(m)
  })
}) -> DEmat

# load groups
grpData <- readRDS(file.path(orthoPath,"splitGroups/goodTreeStats.rds"))
grps <- loadOrthoGrpsArray(file.path(orthoPath,"splitGroups/goodGroups.txt"))

```

```{r loadPAMLresults, echo=FALSE}
if(!file.exists(pamlResFile)){
  PAMLpath <- "/mnt/NOBACKUP/mariansc/share/orthos/PAML"
  pamlResFiles <- dir(PAMLpath,pattern = "grp.*.out$",recursive = T,full.names = T)
  names(pamlResFiles) <- sub(".*/(grp[0-9\\.]+_.*).out","\\1", pamlResFiles,perl=T)
  
  sapply(pamlResFiles,function(outFile){
    txt <- readLines(outFile)
    idx <- grep("lnL",txt)
    if(length(idx)==1){
      return(as.numeric(sub(".*(-[0-9\\.]+).*","\\1",txt[idx],perl=T)))
    } else{
      return(NA_real_)
    }
  }) -> lnL
  
  # make table from the names
  tbl <- readr::read_delim(paste(names(lnL),collapse="\n"),delim="_",col_names=c("grpID","hyp","H"))
  tbl$lnL <- lnL
  
  tblWide <- tidyr::spread(tbl,H,lnL)
  tblWide$pVal <- 1-pchisq(2*(tblWide$H1-tblWide$H0),1)
  pamlRes <- split(tblWide[,colnames(tblWide)!="hyp"],tblWide$hyp)
  
  pamlRes <- lapply(pamlRes,as.data.frame)
  
  saveRDS(pamlRes,pamlResFile)
} else {
  pamlRes <- readRDS(pamlResFile) 
}

# remove NA's in H4a (probably caused by trees where the core clade is on the root and can't be marked)
pamlRes$H4a <- pamlRes$H4a[!is.na(pamlRes$H4a$pVal), ]


# convert to matrix of p-values
pamlMat <- sapply(pamlRes, with, {pVal[match(rownames(expr),grpID)]})

```

### Postive selection
```{r pamlResPlot, echo=FALSE}
# check how many significant there are in each hypothesis
pamlSign <- sapply(pamlRes, with, {sum(pVal < 0.05)})
nPaml <- sapply(pamlRes, nrow)
barplot(pamlSign/nPaml, ylab="proportion of sign. tests (p<0.05)",main="positive selection")
```


### Common responses in sub-set of species

#### Genes responding to cold in all pooids

Which genes are up-regulated in all our five pooids as respons to cold?

Method one: Select genes with p.adj<0.05 and lfc>1 in all species. Ramp and Peak effect tested individually.


```{r echo=FALSE}
sigMat <- lapply(DEmat, with, {pAdj < 0.05 & lfc > 1})

venn(apply(sigMat$peak,2,which))
title(main="significant peak up",sub="p.adj<0.05 & lfc>1")

whichAllpeakUp <- which(apply(sigMat$peak,1,all))

heatmap(DEmat$peak$stat[whichAllpeakUp,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13),
        RowSideColors = ifelse(pamlMat[whichAllpeakUp,"H4a"]<0.05,"red","grey"),
        scale="none", Rowv = NA, Colv = NA, main="Peak up in all (H4a)")

venn(apply(sigMat$ramp,2,which))
title(main="significant ramp up",sub="p.adj<0.05 & lfc>1")

whichAllrampUp <- which(apply(sigMat$ramp,1,all))

heatmap(DEmat$ramp$stat[whichAllrampUp,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13),
        RowSideColors = ifelse(pamlMat[whichAllrampUp,"H4a"]<0.05,"red","grey"),
        scale="none", Rowv = NA, Colv = NA, main="Ramp up in all (H4a)")

```


### Alternative way of plotting expression profiles

```{r defFun, echo=FALSE}
getExpr <- function( spc, seqIDs){
  DE[[spc]]$vst[seqIDs, ,drop=F]
}


plotProfiles <- function(grp, getLty=function(spc,genes){1}){
  if(length(unlist(grp))==0)
    return

  spcsExpr <- names(DE) # defines which species to plot profiles for
  
  layout(matrix(c(1:length(spcsExpr)), 1, length(spcsExpr), byrow = T),
         widths=rep(2,length(spcsExpr)), heights=c(2.5))
  par(mar=c(5, 3, 4, 0))
  TtoX <- c(`T-1`=0,T0=1,T1=2,T3=3,T4=4)

  
  # get same scale for all species
#   ylim <- range(unlist(lapply(spcs,function(spc){expr[[spc]][grp[[spc]],]})))
  ylim <- c(0,15)
  
  for(spc in spcsExpr){
    genes <- grp[[spc]]
    
    expCnt <- getExpr(spc,genes)    
  
    plot(x=NULL,xaxt="n",main=spc,xlab="",
         xlim=range(TtoX), ylim=ylim)
    if(spc==spcsExpr[1]){
      title(ylab="expr. (VST)",line=2)
    }
    axis(1,at=TtoX,labels=names(TtoX)) 
    timePoint = str_extract(colnames(expCnt),"T-?[0-4]")
    cols <- rainbow(length(genes))
    for( i in seq_along(genes) ) {
      points(x=TtoX[timePoint],y=expCnt[i,], col=cols[i])
      exprMean <- tapply(as.numeric(expCnt[i,]),timePoint,mean)
      exprMean <- exprMean[order(TtoX[names(exprMean)])]
      lines(x=TtoX[names(exprMean)],y=exprMean, col=cols[i], lty=getLty(spc,genes[i]))
    }
  }
  # reset graphics parameters
  layout(1)
  par(mar=c(5, 4, 4, 2) + 0.1)
}

plotProfiles2 <- function(grp){
  if(length(unlist(grp))==0)
    return

  spcsExpr <- names(DE) # defines which species to plot profiles for
  
  layout(matrix(c(1:(2*length(spcsExpr))), 2, length(spcsExpr), byrow = F),
         widths=rep(2,length(spcsExpr)), heights=c(1.3,1.3))
  par(mar=c(5, 3, 4, 0))
  
  TtoF <- c(`T-1`="ramp",T0="peak",T1="peak",T3="ramp",T4="ramp")    
  TtoX <- c(`T-1`=0,T0=0,T1=1,T3=1,T4=1)
  FtoCol <- c(ramp="red",peak="blue")
  
  # get same scale for all species
  
  for(spc in spcsExpr){

    genes <- grp[[spc]]
    
    expCnt <- getExpr(spc,genes)    

    timePoint = str_extract(colnames(expCnt),"T-?[0-4]")
    xf <- as.factor(paste0(TtoF[timePoint],TtoX[timePoint]))
    exprMean <- tapply(as.numeric(expCnt),rep(xf,each=nrow(expCnt)),function(lgCnt){
      log2(mean(2^lgCnt))
    })
    
    for( peakRamp in c("peak","ramp")){
      idx <- TtoF[timePoint]==peakRamp
      col <- FtoCol[peakRamp]
      
      if( peakRamp=="peak"){
        ymean <- exprMean[c("peak0","peak1")]
      } else{
        if("ramp0" %in% names(exprMean))
          ymean <- exprMean[c("ramp0","ramp1")]
        else
          ymean <- exprMean[c("peak0","ramp1")]
      }
      
      ylim <- c(-2,2)+mean(ymean)
      
      plot(x=NULL,xaxt="n",main=spc,xlab="",
           xlim=range(TtoX), ylim=ylim)
      title(sub=peakRamp,line=2)

      if(spc==spcsExpr[1]){
        title(ylab="expr. (VST)",line=2)
      }
      axis(1,at=c(0,1),labels=c("ref","cold")) 
      
      for( i in seq_along(genes) ) {
        points(x=TtoX[timePoint][idx],y=expCnt[i,idx], col=col)
      }
      
      lines(x=c(0,1),y=ymean, col=col)
    }
  }
  # reset graphics parameters
  layout(1)
  par(mar=c(5, 4, 4, 2) + 0.1)
}

```

#### profile for grp100339.1 (peak up in all)
```{r echo=FALSE}
plotProfiles2(grps["grp100339.1", ])
```

#### profile for grp101184.1 (ramp up in all)
```{r echo=FALSE}
plotProfiles2(grps["grp101184.1", ])
```


```{r echo=FALSE,eval=FALSE}
#Method two: The average test statistic in each species is greater than 0

lapply(DEmat, with, {apply(ifelse(stat>5,5,stat), 1, function(x){
  t.test(x,alternative="greater")$p.value})
}) -> pValMethod2

dim(DEmat$peak$stat)
range(ifelse(DEmat$peak$stat>5,5,DEmat$peak$stat))
lapply(DEmat, with, {apply(stat, 1, function(x){
  t.test(x,alternative="greater")$p.value})
}) -> pValMethod3

table(pValMethod2$peak<0.05,pValMethod3$peak<0.05)
pValMethod2$peak
t.test(c(5,5,5,5,4),alternative = "greater")

table(p.adjust(pValMethod2$peak,method="BH")<0.05)
```




```{r ttestOnStat, echo=FALSE,cache=TRUE}
apply(DEmat$ramp$stat,1,function(statRow){
  res <- t.test(x = statRow[1:2],y = statRow[3:5],alternative = "greater")
  return(c(pVal=res$p.value, tStat=res$statistic))
}) -> rampCoreVsBasalttest

apply(DEmat$peak$stat,1,function(statRow){
  res <- t.test(x = statRow[1:2],y = statRow[3:5],alternative = "greater")
  return(c(pVal=res$p.value, tStat=res$statistic))
}) -> peakCoreVsBasalttest

# cor.test(rampCoreVsBasalttest["tStat.t", ], peakCoreVsBasalttest["tStat.t", ])
# 
# apply(cbind(statMatrixPeak,statMatrixRamp),1,function(statRow){
#   res <- t.test(x = statRow[c(1:2,6:7)],y = statRow[c(3:5,8:10)],alternative = "greater")
#   return(c(pVal=res$p.value, tStat=res$statistic))
# }) -> comboCoreVsBasalttest
# 
# hist(comboCoreVsBasalttest["pVal",])
# table(p.adjust(comboCoreVsBasalttest["pVal",],method = "fdr")<0.1)


# hist(peakCoreVsBasalttest["pVal",])
# table(p.adjust(peakCoreVsBasalttest["pVal",],method = "fdr")<0.1)
# hist(rampCoreVsBasalttest["pVal",])
# table(p.adjust(rampCoreVsBasalttest["pVal",],method = "fdr")<0.1)

```

### Core specific response


```{r ploth4b, echo=FALSE}

idx <- order(rampCoreVsBasalttest["tStat.t",],decreasing = T)[1:40]

heatmap(DEmat$ramp$stat[idx,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13),
        RowSideColors = ifelse(pamlMat[idx,"H4a"]<0.05,"red","grey"),
        scale="none", Colv = NA,main="Ramp up in core (H4b)")

idx <- order(peakCoreVsBasalttest["tStat.t",],decreasing = T)[1:40]

heatmap(DEmat$peak$stat[idx,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13),
        RowSideColors = ifelse(pamlMat[idx,"H4a"]<0.05,"red","grey"),
        scale="none", Colv = NA,main="Peak up in core (H4b)")

# 
# i=1
# i=i+1
# grpIDs[i]
# plotProfiles2(grp=grps[grpIDs[i],])
# statMatrix[grpIDs[i], ]
# 


# correlate H4b (positive selection in root of core clade) with t-test

# cor(statMatrix)
# sum(tStat < 0.05)
# 
# grpIDsInBoth <- rownames(expr)[rownames(expr) %in% H4b$grpID[!is.na( H4b$pVal )]]
# 
# cor.test(x=tStat[match(grpIDsInBoth, rownames(expr))],
#     y=as.numeric(H4b$pVal[match(grpIDsInBoth,H4b$grpID)] < 0.05))
# range(-log10(H4b$pVal[match(grpIDsInBoth,H4b$grpID)]))
# table(H4b$pVal[match(grpIDsInBoth,H4b$grpID)] == 0)
# 
# table(is.na(-log10(H4b$pVal)[grpIDsInBoth]))
# 
# tStat[rownames(expr)[idx] %in% ]
# 
# grpIDsByH4b <- H4b$grpID[order(H4b$pVal)]
# grpIDsByH4b <- sample(H4b$grpID[order(H4b$pVal)])
# grpIDsTopByH4b <- grpIDsByH4b[grpIDsByH4b %in% rownames(statMatrix)][1:40]
# 
# heatmap(statMatrix[grpIDsTopByH4b,], breaks=c(-20,seq(-5.5,5.5,by=1),20),
#         col=colorRampPalette(c("green","black","red"))(13),
#         RowSideColors = ifelse(H4b[grpIDsTopByH4b,"pVal"]<0.05,"red","grey"),
#         scale="none", Rowv = NA)

```


