---
title: "Analyse Replicate Bias"
output: html_document
---

### Goal of analysis

Each species in our dataset has a different number of replicates which affects the number of significant DEG detected. In this analysis I shall visualise the extent of this technical bias and try to find a way to make fair comparisons between species despite it.

```{r loadLibraries, echo=F, message=F}
library(DESeq2)
library(stringr)

source("~/GRewd/pipeline/R/orthoGrpTools.R")
```


```{r loadData, echo=FALSE, cache=T}
# define paths
pipelineOutDir <- "/mnt/NOBACKUP/mariansc/share"
tblDir <- file.path(pipelineOutDir,"orthos/exprTbls")
dataPath <- "~/GRewd/pipeline/analysis/replicateBiasAnalysis_data"

# load DE results
load("/mnt/NOBACKUP/mariansc/share/orthos/DESeq/DE.RData")

grpData <- readRDS("/mnt/NOBACKUP/mariansc/share/orthos/splitGroups/goodTreeStats.rds")
grps <- loadOrthoGrpsArray("/mnt/NOBACKUP/mariansc/share/orthos/splitGroups/goodGroups.txt")

# select high confidence groups (1:1 with correct topology)
grpIdx <- with(grpData, hasAllDenovo & !hasParalogs & allSpcAreClans & isCoreClan)

# make matrix out of DE results
lapply(c(ramp="resRamp",peak="resPeak"),function(eff){
  lapply(c(Qval="padj",FC="log2FoldChange",Pval="pvalue",mean="baseMean",stat="stat"),function(stat){
    sapply(names(DE), function(spc){
      DE[[spc]][[eff]][unlist(grps[grpIdx,spc]),stat]
    }) 
  })
}) -> resArray

```


### Some stats

```{r showTable, echo=FALSE, results='asis'}
nReps <- sapply(DE,function(x){ncol(x$vst)})
nGenes <- sapply(DE,function(x){nrow(x$vst)})

nDEGs <- sapply(DE,function(x){sum(x$resPeak[ ,"pvalue"]<0.05 | 
                                     x$resRamp[ ,"pvalue"]<0.05 ,na.rm = T)})

knitr::kable(data.frame(nReps=nReps,
                        `T-1`= sapply(DE,function(x){sum(grepl("T-1",colnames(x$vst)))}),
                        `T0`= sapply(DE,function(x){sum(grepl("T0",colnames(x$vst)))}),
                        `T1`= sapply(DE,function(x){sum(grepl("T1",colnames(x$vst)))}),
                        `T3`= sapply(DE,function(x){sum(grepl("T3",colnames(x$vst)))}),
                        `T4`= sapply(DE,function(x){sum(grepl("T4",colnames(x$vst)))}),
                        nGenes=nGenes,nDEGs=nDEGs,
                        row.names = names(DE)))

barplot(rbind(nGenes=nGenes,nDEGs=nDEGs), beside=T,col=c("brown","wheat"))
legend("topleft",legend=c("nGenes","nDEGs"), fill=c("brown","wheat"))
```

```{r plotRepsVsDEGs, echo=FALSE}

plot(nReps,nDEGs/nGenes,pch=NA)
text(nReps,nDEGs/nGenes,labels = names(DE))
title(sub="DE if P<0.05 in either ramp or peak")

```

The following plots show the distribution of log fold change values of the DEGs for either ramp or peak (P<0.05). The area under the curve equals the number of DEGs.

```{r DEGdensityplot, echo=FALSE}

lapply(list(ramp="resRamp",peak="resPeak"),function(eff){
  lapply(DE,function(x){
    res <- x[[eff]][ ,"pvalue"]<0.05
    res[is.na(res)] <- F
    return(res)
  })
}) -> DEGs

spcs <- setNames(names(DE),names(DE))
spcColors <- c(BrDi="#00A0B0", HoVu="#6A4A3C", MeNu1="#CC333F",
               MeNu2="#EB9891", NaSt="#EDC951", StLa="#2ECC71")

plot(NA, xlim=c(-6,6),ylim=c(0,5300),xlab="lfc",ylab="Density*nDEGs", 
     main="DEG ramp")
for( spc in spcs ){
  d <- density(DE[[spc]]$resRamp[ DEGs$ramp[[spc]],"log2FoldChange"],bw=0.1)
  
  lines(d$x,d$y*sum(DEGs$ramp[[spc]]),
        col=spcColors[spc],lwd=2)  
}
legend("topleft",legend = names(spcColors), col=spcColors,lwd=2)

plot(NA, xlim=c(-6,6),ylim=c(0,5300),xlab="lfc",ylab="Density*nDEGs", 
     main="DEG peak")
for( spc in spcs ){
  d <- density(DE[[spc]]$resPeak[ DEGs$peak[[spc]],"log2FoldChange"],bw=0.1)
  
  lines(d$x,d$y*sum(DEGs$peak[[spc]]),
        col=spcColors[spc],lwd=2)  
}
legend("topleft",legend = names(spcColors), col=spcColors,lwd=2)

```

### High confidence 1:1:1 ortho groups

Because the number of genes varies between species I want to compare the species using the `r nrow(resArray$peak$Qval)` high confidence groups (so we know we are comparing the same set of genes).

```{r highConGroups, echo=FALSE}
n <- nrow(resArray$peak$Qval) # number of 1:1:1 groups

barplot( rbind(apply(resArray$peak$Pval < 0.05,2,sum,na.rm = T),
               apply(resArray$ramp$Pval < 0.05,2,sum,na.rm = T)),
         beside=T, main="DEGs in high confidence 1:1:1 groups",
         ylab="DEGs", sub="(P < 0.05)",
         col=c("wheat1","wheat3"))
legend("topleft",legend=c("peak","ramp"),fill=c("wheat1","wheat3"))

nDEGs11 <- apply(resArray$peak$Pval < 0.05 | resArray$ramp$Pval < 0.05,2,sum,na.rm = T)
plot(nReps,nDEGs11/n,pch=NA, ylab="proportion of DEGs",
     main="DEGs in high confidence 1:1:1 groups")
text(nReps,nDEGs11/n,labels = names(DE))
title(sub="DE if P<0.05 in either ramp or peak")

```

### Conclusion 1

There is a clear tendancy that species with few samples have fewer DEGs

### Removing samples

To see if the relatively low number of significant DEGs in StLa is caused by the low number of samples, we can discard samples from the other species so that they get the same experiment design.

```{r calcReduceDE, echo=FALSE}

if(file.exists(file.path(dataPath,"reducedDE.RData"))){
  load(file.path(dataPath,"reducedDE.RData"))
} else {
  
  # set some useful constants:
  spcs <- c("HoVu","MeNu1","NaSt") # only need these..
  spcs <- setNames(spcs,spcs)
  
  TtoX <- c(`T-1`=0,T0=1,T1=2,T3=3,T4=4)
  TtoF <- c(`T-1`="ramp0",T0="peak0",T1="peak1",T3="ramp1",T4="ramp1")
  
  # read the full expression count tables:
  
  
  files <- dir(tblDir,pattern=paste0("(",paste(spcs,collapse="|"),")_expected_countTbl.txt"),full.names = T)
  exprCnt <- lapply(setNames(files,str_extract(basename(files),"^[^_]+")),
                    read.table, stringsAsFactors = F)
  
  
  # use a random HoVu.T1.ind for HoVu.T1.mix
  exprCnt$HoVu$HoVu.T1.mix.1 <- exprCnt$HoVu$HoVu.T1.ind.2
  
  
  # get the samples in corresponding to StLa samples
  reducedSampleIDs <- c(sub("StLa","NaSt1",colnames(DE$StLa$vst)),
                        sub("StLa","MeNu1",colnames(DE$StLa$vst)),
                        sub("StLa","HoVu",colnames(DE$StLa$vst)))
  
  # for each species:
  lapply(exprCnt,function(countData){
  
    # discard samples
    countData <- countData[names(countData) %in% reducedSampleIDs]
    
    colData <- as.data.frame(str_split_fixed(names(countData),"\\.",4)[ ,1:3])
    names(colData) <- c("spcPop","timePoint","mix")
    colData$Tf <- as.factor(TtoF[as.character(colData$timePoint)])
    
    
    dds <- DESeqDataSetFromMatrix(countData = round(countData),
                                  colData = colData,
                                  design = formula(~ Tf ))
    
    dds <- DESeq(dds)
    
    # contrast tests
    if( "Tframp0"  %in% resultsNames(dds)){
      resRamp <- results(dds, contrast=c("Tf","ramp1","ramp0"))
    } else { # T-1 (ramp0) is missing in StLa... use T0 instead
      resRamp <- results(dds, contrast=c("Tf","ramp1","peak0"))    
    }
    resPeak <- results(dds, contrast=c("Tf","peak1","peak0"))
    
    return(list(resRamp=resRamp,resPeak=resPeak))
  }) -> reducedDE
  
  dir.create(dataPath)
  save(reducedDE,file=file.path(dataPath,"reducedDE.RData"))
}

```

```{r plot11redRepVsDEGs, echo=FALSE}
lapply(c(ramp="resRamp",peak="resPeak"),function(eff){
  lapply(c(Pval="pvalue"),function(stat){
    sapply(names(reducedDE), function(spc){
      reducedDE[[spc]][[eff]][unlist(grps[grpIdx,spc]),stat]
    }) 
  })
}) -> resArrayReduced


colnames(resArrayReduced$ramp$Pval) <- paste0(colnames(resArrayReduced$ramp$Pval),"_r")
colnames(resArrayReduced$peak$Pval) <- paste0(colnames(resArrayReduced$peak$Pval),"_r")


nDEGs11red <- apply(cbind(resArray$peak$Pval,resArrayReduced$peak$Pval) < 0.05 | 
                      cbind(resArray$ramp$Pval,resArrayReduced$ramp$Pval) < 0.05,
                    2,sum,na.rm = T)
nRepsRed <- c(nReps,7,7,7)
plot(nRepsRed,nDEGs11red/n,pch=NA, ylab="proportion of DEGs",
     main="DEGs in high confidence 1:1:1 groups")
text(nRepsRed,nDEGs11red/n,labels = colnames(cbind(resArray$peak$Pval,resArrayReduced$peak$Pval)))
title(sub="DE if P<0.05 in either ramp or peak")


```

When removing a samples so that each species have the same design HoVu and MeNu1 end up with the same number of DEGs. The number of DEGs in NaSt is not reduced nearly as much as HoVu and MeNu1. One possible explanation could be that NaSt was sampled from 3 populations and the reduced set only includes one of those, therefore removing the variance between population. In the plots below we can se that the DEGs in NaSt is from mainly in the ramp effect.


```{r plotDensityReducedSamples, echo=FALSE}

reducedDE$StLa <- DE$StLa # copy StLa for convenience


lapply(list(ramp="resRamp",peak="resPeak"),function(eff){
  lapply(reducedDE,function(x){
    res <- x[[eff]][ ,"pvalue"]<0.05
    res[is.na(res)] <- F
    return(res)
  })
}) -> DEGs

spcs <- setNames(names(reducedDE),names(reducedDE))

plot(NA, xlim=c(-6,6),ylim=c(0,5300),xlab="lfc",ylab="Density*nDEGs", 
     main="DEG ramp (fewer samples)")
for( spc in spcs ){
  d <- density(reducedDE[[spc]]$resRamp[ DEGs$ramp[[spc]],"log2FoldChange"],bw=0.1)
  
  lines(d$x,d$y*sum(DEGs$ramp[[spc]]),
        col=spcColors[spc],lwd=2)  
}
legend("topleft",legend = spcs, col=spcColors[spcs],lwd=2)

plot(NA, xlim=c(-6,6),ylim=c(0,5300),xlab="lfc",ylab="Density*nDEGs", 
     main="DEG peak (fewer samples)")
for( spc in spcs ){
  d <- density(reducedDE[[spc]]$resPeak[ DEGs$peak[[spc]],"log2FoldChange"],bw=0.1)
  
  lines(d$x,d$y*sum(DEGs$peak[[spc]]),
        col=spcColors[spc],lwd=2)  
}
legend("topleft",legend = spcs, col=spcColors[spcs],lwd=2)
```


### Conclusion 2

When taking into account the number of replicates there are still fewer DEGs in StLa, which indicates that it is a biological effect.

### Assuming common variance across species

Using the 1:1:1 groups it is possible to let DESeq use all species to estimate variance.

```{r calcCommonVarDE, echo=FALSE}
if(file.exists(file.path(dataPath,"commonVarDE.RData"))){
  load(file.path(dataPath,"commonVarDE.RData"))
} else {
  
  # set some useful constants:
  spcs <- names(DE) 
  spcs <- setNames(spcs,spcs)
  
  TtoX <- c(`T-1`=0,T0=1,T1=2,T3=3,T4=4)
  TtoF <- c(`T-1`="ramp0",T0="peak0",T1="peak1",T3="ramp1",T4="ramp1")
  
  # read the full expression count tables:
  
  
  files <- dir(tblDir,pattern=paste0("(",paste(spcs,collapse="|"),")_expected_countTbl.txt"),full.names = T)
  exprCnt <- lapply(setNames(files,str_extract(basename(files),"^[^_]+")),
                    read.table, stringsAsFactors = F)
  
  
  exprCnt111 <- Reduce(cbind,
                       lapply(spcs,function(spc){
                         exprCnt[[spc]][unlist(grps[grpIdx,spc]), ]
                        }))
  
  
  # fix the T-1 name
  names(exprCnt111) <- sub("T\\.1","T-1",names(exprCnt111))
  
  
  colData <- as.data.frame(str_split_fixed(names(exprCnt111),"\\.",4)[ ,1:3])
  names(colData) <- c("spcPop","timePoint","mix")
  colData$spc <- str_extract(colData$spcPop,paste(spcs,collapse="|"))
  colData$Tf <- as.factor(TtoF[as.character(colData$timePoint)])
  colData$TfSpc <- factor(paste0(colData$Tf,colData$spc))

  dds <- DESeqDataSetFromMatrix(countData = round(exprCnt111),
                                  colData = colData,
                                  design = formula(~ TfSpc ))
    
  dds <- DESeq(dds)
  
  
  # contrast tests fo each species
  lapply(spcs,function(spc){
    if( spc != "StLa"){
      resRamp <- results(dds, contrast=c("TfSpc",paste0("ramp1",spc),paste0("ramp0",spc)))
    } else { # T-1 (ramp0) is missing in StLa... use T0 instead
      resRamp <- results(dds, contrast=c("TfSpc",paste0("ramp1",spc),paste0("peak0",spc)))    
    }
    resPeak <- results(dds, contrast=c("TfSpc",paste0("peak1",spc),paste0("peak0",spc)))
    
    return(list(resRamp=resRamp,resPeak=resPeak))
  }) -> commonVarDE

  save(commonVarDE,file=file.path(dataPath,"commonVarDE.RData"))
}

```



```{r plotDEGsCommonVar, echo=F}
barplot( rbind(sapply(commonVarDE,function(x){ sum(x$resPeak[, "pvalue"]<0.05,na.rm=F)}),
               sapply(commonVarDE,function(x){ sum(x$resRamp[, "pvalue"]<0.05,na.rm=F)})),
         beside=T, main="common variance DEGs",
         ylab="DEGs", sub="(P < 0.05)",
         col=c("wheat1","wheat3"))
legend("bottomleft",legend=c("peak","ramp"),fill=c("wheat1","wheat3"))
```


```{r plotDensityCommonVar, echo=FALSE}


lapply(list(ramp="resRamp",peak="resPeak"),function(eff){
  lapply(commonVarDE,function(x){
    res <- x[[eff]][ ,"pvalue"]<0.05
    res[is.na(res)] <- F
    return(res)
  })
}) -> DEGs

spcs <- setNames(names(commonVarDE),names(commonVarDE))

plot(NA, xlim=c(-6,6),ylim=c(0,1100),xlab="lfc",ylab="Density*nDEGs", 
     main="DEG ramp (common variance)")
for( spc in spcs ){
  d <- density(commonVarDE[[spc]]$resRamp[ DEGs$ramp[[spc]],"log2FoldChange"],bw=0.1)
  
  lines(d$x,d$y*sum(DEGs$ramp[[spc]]),
        col=spcColors[spc],lwd=2)  
}
legend("topleft",legend = spcs, col=spcColors[spcs],lwd=2)

plot(NA, xlim=c(-6,6),ylim=c(0,1100),xlab="lfc",ylab="Density*nDEGs", 
     main="DEG peak (common variance)")
for( spc in spcs ){
  d <- density(commonVarDE[[spc]]$resPeak[ DEGs$peak[[spc]],"log2FoldChange"],bw=0.1)
  
  lines(d$x,d$y*sum(DEGs$peak[[spc]]),
        col=spcColors[spc],lwd=2)  
}
legend("topleft",legend = spcs, col=spcColors[spcs],lwd=2)
```
