---
title: "Broadly conserved response plot"
output: html_document
---

### Goal of analysis

The point is to illustrate that even though there are seemingly few single genes with conserved cold response across the pooids, there is a clear correlation when considering the average of many genes. To do this I have plotted the sliding window (windowsize=500 genes) mean lfc (y-axis) against the rank of lfc (x-axis). Only orthogroups expressed in all species are included (4981 orthogroups).

```{r loadData, echo=FALSE}
DEmat <- readRDS("~/GRewd/pipeline/data/DEmat.RDS")  # from: superGeneModel.Rmd
vst <- readRDS("~/GRewd/pipeline/data/VSTnew.RDS")

spcs <- colnames(DEmat$peak$lfc)
names(spcs) <- spcs

filterLowExpr <- apply(vst>3,1,all)
```



```{r plotMatrix, echo=F, fig.height=9,fig.width=9}
slideMean<-function(x,windowsize=3,slide=2){
  idx1<-seq(1,length(x)-windowsize+1,by=slide);
  idx2<-seq(windowsize,length(x),by=slide);
  cx <- c(0,cumsum(x));
  return((cx[idx2+1]-cx[idx1])/windowsize);
}

m <- matrix(1:25, 5, 5, byrow = TRUE)
layout(rbind(cbind(m,m+25),cbind(m+50,m+75)))
par(mar=c(0,0,0,0))
for(eff1 in c("peak","ramp")){
  for(eff2 in c("peak","ramp")){
    for(spc1 in spcs){
      for(spc2 in spcs){
        plot(NULL,type="n",ylim=c(-1,1),xlim=c(0,1),
             axes = F,xlab=NA,ylab=NA,lwd=2)

        y <- DEmat[[eff1]]$lfc[filterLowExpr,spc1]
#           y <- rank(DEmat[[eff1]]$lfc[filterLowExpr,spc1])
        idx <- order(DEmat[[eff2]]$lfc[filterLowExpr,spc2])
        yMean <- slideMean(y[idx], windowsize=500,slide=50)
#           plot(seq_along(yMean),yMean,type="l",ylim=c(1000,3981),
#                axes = F,xlab=NA,ylab=NA,lwd=2)

        if(spc1==spc2 & eff1 == eff2){
          text(0.5,-0.4,labels = spc1)
          text(0.5,0.4,labels = c(peak="Short term",ramp="Long term")[eff1])
          box()
        } else {
          n=5
          tapply(idx,cut(seq_along(idx),breaks = n,labels = 1:n), function(i){
            cor.test(y[i],DEmat[[eff2]]$lfc[filterLowExpr,spc2][i],method="spearman")$p.value          
          }) -> corPval
  
          cols <- colorRampPalette(c("white","#8899EE"))(6)
          idxCorPval <- floor(-log10(corPval))+1
          idxCorPval[idxCorPval>length(cols)] <- length(cols)
          for(i in 1:n){
            rect(seq(0,1,length.out = n+1)[i],-1,seq(0,1,length.out = n+1)[i+1],1,
                 col = cols[idxCorPval[i]],border = NA)
          }
        }


        abline(h=-0.5,lty=2,col="lightgray")
        abline(h=0.5,lty=2,col="lightgray")
        abline(h=0,lty=1, col="lightgray")
        lines(seq(0,1,length.out = length(yMean)),yMean,lwd=2)
             
      }  
    }
  }
}
par(mar=c(5.1,4.1,4.1,2.1))
layout(1)

```

```{r}
DEdiurnal <- readRDS("~/GRewd/pipeline/data/DEdiurnal.RDS")

m <- matrix(1:16, 4, 4, byrow = TRUE)
layout(m)
par(mar=c(0,0,0,0))
for(spc1 in spcs[-4]){
  for(spc2 in spcs[-4]){
    if(spc1==spc2 & eff1 == eff2){
      plot(NULL,xlim=c(-1,1),ylim=c(-2.5,2.5),axes = F,xlab=NA,ylab=NA)
      text(0,0,labels = spc1)
      box()
    } else {
      y <- DEdiurnal[[spc1]]$log2FoldChange[filterLowExpr]
#           y <- rank(DEmat[[eff1]]$lfc[filterLowExpr,spc1])
      idx <- order(DEdiurnal[[spc2]]$log2FoldChange[filterLowExpr])
      yMean <- slideMean(y[idx], windowsize=500,slide=50)
      plot(NULL,type="n",ylim=c(-1,1),xlim=c(0,1),
           axes = F,xlab=NA,ylab=NA,lwd=2)
#           plot(seq_along(yMean),yMean,type="l",ylim=c(1000,3981),
#                axes = F,xlab=NA,ylab=NA,lwd=2)

      n=5
      tapply(idx,cut(seq_along(idx),breaks = n,labels = 1:n), function(i){
        cor.test(y[i],DEdiurnal[[spc2]]$log2FoldChange[filterLowExpr][i],method="spearman")$p.value          
      }) -> corPval

      cols <- colorRampPalette(c("white","#8899EE"))(6)
      idxCorPval <- floor(-log10(corPval))+1
      idxCorPval[idxCorPval>length(cols)] <- length(cols)
      for(i in 1:n){
        rect(seq(0,1,length.out = n+1)[i],-1,seq(0,1,length.out = n+1)[i+1],1,
             col = cols[idxCorPval[i]],border = NA)
      }

      abline(h=-0.5,lty=2,col="lightgray")
      abline(h=0.5,lty=2,col="lightgray")
      abline(h=0,lty=1, col="lightgray")
      lines(seq(0,1,length.out = length(yMean)),yMean,lwd=2)
           
    }
  }  
}
par(mar=c(5.1,4.1,4.1,2.1))
layout(1)

```

