---
title: "Supergene Clustering"
output: html_document
---


```{r loadLibraries, echo=FALSE, message=FALSE}
library(DESeq2)
library(WGCNA)


source("~/GRewd/pipeline/R/orthoGrpTools.R")
```

```{r loadData, echo=FALSE}
orthoPath <- "/mnt/NOBACKUP/mariansc/share/orthos"


modulesVST.File <- "~/GRewd/pipeline/data/modulesVST.RDS"
modulesDE.File <- "~/GRewd/pipeline/data/modulesDE.RDS"
pamlResFile <- "~/GRewd/pipeline/data/pamlRes.RDS"
DEnewFile <- "~/GRewd/pipeline/data/DEnew.RDS"
VSTnewFile <- "~/GRewd/pipeline/data/VSTnew.RDS"
superExprMatrixFile <- "~/GRewd/pipeline/data/superExprMatrix.RDS"

# load DE
load(file.path(orthoPath,"DESeq/DE.RData"))

# load superGenes expression
expr <- readRDS(file = superExprMatrixFile)  
# load DE based and vst based on superGenes
DEnew <- readRDS(DEnewFile)  
vst <- readRDS(VSTnewFile)


# convert DE results to matrix format
lapply( list( peak="resPeak", ramp="resRamp"), function( peakOrRamp ){
  lapply( list( pVal="pvalue", pAdj="padj", stat="stat",lfc="log2FoldChange"), function( stat ){
    m <- sapply(DEnew, function(DEspc){DEspc[[peakOrRamp]][[stat]]})
    rownames(m) <- rownames(expr)
    return(m)
  })
}) -> DEmat

# load groups
grpData <- readRDS(file.path(orthoPath,"splitGroups/goodTreeStats.rds"))
grps <- loadOrthoGrpsArray(file.path(orthoPath,"splitGroups/goodGroups.txt"))

```

```{r loadPAMLresults, echo=FALSE}
pamlRes <- readRDS(pamlResFile) 

# remove NA's in H4a (probably caused by trees where the core clade is on the root and can't be marked)
pamlRes$H4a <- pamlRes$H4a[!is.na(pamlRes$H4a$pVal), ]


# convert to matrix of p-values
pamlMat <- sapply(pamlRes, with, {pVal[match(rownames(expr),grpID)]})

```


### Find gene clusters in the variance stabilized supergene matrix using WGCNA

```{r findModules, echo=FALSE}
if( !file.exists(modulesVST.File)){
  modulesVST <- blockwiseModules(datExpr = t(vst), power=10, minModuleSize=40, minAbsSplitHeight=0.1)
  saveRDS(modulesVST,modulesVST.File)
} else{
  modulesVST <- readRDS(modulesVST.File)  
}
```

```{r plotModulesVST, echo=FALSE}
#plot(table(modules$colors))
m <- as.matrix(modulesVST$MEs)
rownames(m) <- colnames(vst)
heatmap(m,scale="none")
```

### Test for over-representation of positive selection in modules

```{r testPhyperVST, echo=FALSE}

sapply(unique(modulesVST$colors), function(mod){
  sapply(colnames(pamlMat), function(hyp){
    idx <- !is.na(pamlMat[ ,hyp])
    phyper(q = sum(modulesVST$colors[idx]==mod & pamlMat[idx,hyp] < 0.05),
           m = sum(pamlMat[idx,hyp] < 0.05),
           n = sum(pamlMat[idx,hyp] >= 0.05),
           k = sum(modulesVST$colors[idx]==mod),
           lower.tail=F)  
  })
}) -> pHyp.VST

```

```{r}
table(pHyp.VST<0.05)
min(pHyp.VST)
table(p.adjust(pHyp.VST,method = "BH")<0.05)
```


### Find gene clusters in the DE stat matrix using WGCNA

```{r findModulesDE, echo=FALSE, cache=TRUE}
statMat <- cbind(DEmat$peak$stat,DEmat$ramp$stat)
colnames(statMat) <- paste0(rep(c("peak","ramp"),each=5),colnames(statMat))

if( !file.exists(modulesDE.File)){
  modulesDE <- blockwiseModules(datExpr = t(statMat), power=10, minModuleSize=40, 
                                minAbsSplitHeight=0.1)
  saveRDS(modulesDE,modulesDE.File)
} else{
  modulesDE <- readRDS(modulesDE.File)  
}
```


```{r plotModulesDE, echo=FALSE}
#plot(table(modulesDE$colors))
m <- as.matrix(modulesDE$MEs)
rownames(m) <- colnames(statMat)
heatmap(t(m),scale="none", breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13))

```


The eigengenes seems to be very faint, and the reason might be that both positive and negative correlation is clustered together. Here is the one of the modules:

```{r plotBrownModule, echo=FALSE}
ColSideColors <- as.character(cut(modulesDE$MEs$MEbrown,breaks=c(-20,seq(-5.5,5.5,by=1),20),labels = colorRampPalette(c("green","black","red"))(13)))
heatmap(statMat[modulesDE$colors == "brown", ],scale="none",ColSideColors = ColSideColors,
        breaks=c(-20,seq(-5.5,5.5,by=1),20),
        col=colorRampPalette(c("green","black","red"))(13))
```


### Test for over-representation of positive selection in modules

```{r testPhyperDE, echo=FALSE}

sapply(unique(modulesDE$colors), function(mod){
  sapply(colnames(pamlMat), function(hyp){
    idx <- !is.na(pamlMat[ ,hyp])
    phyper(q = sum(modulesDE$colors[idx]==mod & pamlMat[idx,hyp] < 0.05),
           m = sum(pamlMat[idx,hyp] < 0.05),
           n = sum(pamlMat[idx,hyp] >= 0.05),
           k = sum(modulesDE$colors[idx]==mod),
           lower.tail=F)  
  })
}) -> pHyp.DE

```

```{r}
table(pHyp.DE<0.05)
min(pHyp.DE)
table(p.adjust(pHyp.DE,method = "BH")<0.05)
```

