---
title: "treeAnalysis5 re-analysis of split trees"
output: html_document
---

```{r loadLibraries,echo=FALSE}
library(ape)
library(phangorn)
source("~/GRewd/pipeline/R/orthoGrpTools.R")
```


```{r loadData, echo=FALSE, cache=TRUE}


myTree <- read.tree(text = "(((((((Bd_R:1,BrDi:1):1,(Hv_R:1,HoVu:1):1):1,(MeNu1:1,MeNu2:1):2):1,StLa:4):1,NaSt:5):1,Os_R:6):1,(Zm_R:2,Sb_R:2):5):0;")


treePath1 <- "/mnt/NOBACKUP/mariansc/share/orthos/treesNuc"
treeFiles1 <- dir(treePath1,pattern="\\.tree$")
treeFiles1 <- setNames(file.path(treePath1,treeFiles1),
                stringr::str_extract(treeFiles1,"grp[0-9]+"))

treePath2 <- "/mnt/NOBACKUP/mariansc/share/orthos/splitOrthos/treesNuc"
treeFiles2 <- dir(treePath2,pattern="\\.tree$")
treeFiles2 <- setNames(file.path(treePath2,treeFiles2),
                stringr::str_extract(treeFiles2,"grp[0-9]+(\\.[0-9]+)?"))


grps <- loadOrthoGrpsArray(orthoGrpFile = "/mnt/NOBACKUP/mariansc/share/orthos/splitOrthos/clanFinder/splitGroups.txt")

origGrpNames <- unique(sub("\\.[0-9]+$","",rownames(grps)))
origGrpTrees <- lapply(treeFiles1[origGrpNames], read.tree)
splitTrees <- lapply(treeFiles2, read.tree)

# get number of genes per species per group
lens <- unlist(lapply(grps,length))
dim(lens) <- dim(grps)
dimnames(lens) <- dimnames(grps)

isOut <- grepl("(Os|Zm|Sb)_R",colnames(lens))

```

```{r defFuns, echo=FALSE}
# Check if tips (given by logical vector isInGrp) form a clan in the tree
isClan <- function(tree, isInGrp){
  clans = getClans(tree)
  any(apply(clans,1,function(clan){
      all(clan==isInGrp)
  }))
}

treePlot <- function(tree, ...){
  plot.phylo(tree,show.node.label = T, 
             tip.color = ifelse(grepl("(Os|Zm|Sb)_R",tree$tip.label),"red",
                                ifelse(grepl("^(H|B)",tree$tip.label),"green","blue")),
             ...)
}

treePlotU <- function(tree, ...){
  plot.phylo(tree,show.tip.label = F,type="u",...)
  tiplabels(pch=20,col=ifelse(grepl("(Os|Zm|Sb)_R",tree$tip.label),"red",
                                  ifelse(grepl("^(H|B)",tree$tip.label),"green","blue")))
}
```

### How many trees have the same topology after re-estimation of ML-phylogen?

There are totally `r length(splitTrees)` trees generated by splitting up `r length(origGrpTrees)` trees.

```{r hasSameTopology, echo=FALSE}

# drop tips from original trees to re-create the
lapply(setNames(names(splitTrees),names(splitTrees)), function(splitGrp){
  origTree <- origGrpTrees[[sub("\\.[0-9]+$","",splitGrp)]]
  splitTree <- splitTrees[[splitGrp]]
  tree <- drop.tip(origTree, origTree$tip.label[!(origTree$tip.label %in% splitTree$tip.label)])
  unroot(tree)
}) -> splitOrigTrees

sapply(names(splitTrees), function(splitGrp){
  ape::all.equal.phylo(splitTrees[[splitGrp]], splitOrigTrees[[splitGrp]], use.edge.length = F)
}) -> isTreeEqual

plot(table(size=sapply(splitTrees,function(tree){length(tree$tip.label)}),isTreeEqual),col=1:2,
     main="How many trees have equal topology?")

```

Totally `r sum(isTreeEqual)` trees have the same topology after re-estimating the phylogeny for just the split sub-tree. Bigger trees are less likely to retain the topology and one of the causes for this is the unclear relationship between paralogs which we don't really care about. 

Let's look at the key topological features and see how they are affected:

* The core species form a clan
* Each species form a separate clan (the pairs MeNu1/MeNu2, HoVu/Hv_R and BrDi/Bd_R are treated as single species)

```{r keyTopoFeaturesPlot, echo=FALSE}

# species
spcs <- c("MeNu1","MeNu2","NaSt","StLa","HoVu","BrDi","Hv_R","Bd_R")

# has core split
isCoreClan <- function(trees){
  sapply(trees,function(tree){
    isCore <- grepl("^(H|B)",tree$tip.label)
    if( sum(isCore) < 2 | sum(!isCore) < 2){
      return(NA) # not informative
    } else {
      return(isClan(tree,isCore))
    }
  })
}

isCoreClan1 <- isCoreClan(splitOrigTrees)
isCoreClan2 <- isCoreClan(splitTrees)

table(isCoreClan1,isCoreClan2, useNA = "always")


areSpcsClans <- function(trees){
  sapply(c("H","B","MeNu","NaSt","StLa"), function(spc){
    sapply(trees,function(tree){
      isInGrp <- grepl(paste0("^",spc),tree$tip.label)
      if( sum(isInGrp) < 2){
        return(NA) # return NA if tip not existing or there is only one tip of this spc
      } else {
        return(isClan(tree,isInGrp))      
      }
    })
  })
}

areSpcsClans1 <- areSpcsClans(splitOrigTrees)
areSpcsClans2 <- areSpcsClans(splitTrees)

table(areSpcsClans1,areSpcsClans2,useNA = "always")

allSpcAreClans1 <- apply(areSpcsClans1,1,function(x){ all(x[!is.na(x)])})
allSpcAreClans2 <- apply(areSpcsClans2,1,function(x){ all(x[!is.na(x)])})

table(allSpcAreClans1,allSpcAreClans2)

goodTopo1 <- allSpcAreClans1 & ifelse(is.na(isCoreClan1),T,isCoreClan1)
goodTopo2 <- allSpcAreClans2 & ifelse(is.na(isCoreClan2),T,isCoreClan2)

table(goodTopo1, goodTopo2)

```

About half of the re-estimated gene-trees have different topology from the original trees. If we look at each species specific branch there is only about 1% disagreement whether they form clans or not. If we look at the core/basal split there is about 10% disagreement. Where there is disagreement it is more often the original tree that fits the species tree.
