---
title: "treeAnalysis4 X:X ortholog trees"
output: html_document
---

```{r loadLibraries,echo=FALSE}
library(ape)
library(phangorn)
source("~/GRewd/pipeline/R/orthoGrpTools.R")
```


```{r loadData, echo=FALSE, cache=TRUE}


myTree <- read.tree(text = "(((((((Bd_R:1,BrDi:1):1,(Hv_R:1,HoVu:1):1):1,(MeNu1:1,MeNu2:1):2):1,StLa:4):1,NaSt:5):1,Os_R:6):1,(Zm_R:2,Sb_R:2):5):0;")


treePath <- "/mnt/NOBACKUP/mariansc/share/orthos/treesNuc"
treeFiles <- dir(treePath,pattern="\\.tree$")
treeFiles <- setNames(file.path(treePath,treeFiles),
                stringr::str_extract(treeFiles,"grp[0-9]+"))

allTrees <- lapply(treeFiles, read.tree)


grps <- loadOrthoGrpsArray(orthoGrpFile = "/mnt/NOBACKUP/mariansc/share/orthos/orthoMCL/groups.txt")

# get number of genes per species per group
lens <- unlist(lapply(grps,length))
dim(lens) <- dim(grps)
dimnames(lens) <- dimnames(grps)

isOut <- grepl("(Os|Zm|Sb)_R",colnames(lens))

```

```{r defFuns, echo=FALSE}

treePlot <- function(tree, ...){
  plot.phylo(tree,show.node.label = T, 
             tip.color = ifelse(grepl("(Os|Zm|Sb)_R",tree$tip.label),"red",
                                ifelse(grepl("^(H|B)",tree$tip.label),"green","blue")),
             ...)
}

treePlotU <- function(tree, ...){
  plot.phylo(tree,show.tip.label = F,type="u",...)
  tiplabels(pch=20,col=ifelse(grepl("(Os|Zm|Sb)_R",tree$tip.label),"red",
                                  ifelse(grepl("^(H|B)",tree$tip.label),"green","blue")))
}
```

### first some stats on the ortholog groups

There are totally `r length(allTrees)` trees that were generated based on ortholog groups with at least 5 sequences. A few (>1%) trees failed to be generated.

```{r someStats, echo=FALSE}

nSpcs <- rowSums(lens[names(allTrees), ]>0)
nGenes <- rowSums(lens[names(allTrees), ])


barplot(table(nSpcs),ylab = "Number of ortholog groups",
        xlab="Number of species in ortholog group")

boxplot(split(nGenes,nSpcs),log = "y",xlab="Number of species in ortholog group",
        ylab="Number of genes in ortholog group")

nOutSpcs <- rowSums(lens[names(allTrees), c("Os_R","Sb_R","Zm_R")]>0)
nCoreSpcs <- rowSums(lens[names(allTrees), c("Bd_R","BrDi","HoVu","Hv_R")]>0)
nBasalSpcs <- rowSums(lens[names(allTrees), c("MeNu1","MeNu2","NaSt","StLa")]>0)

hasOrthoInEach <- (nOutSpcs>0)  & (nCoreSpcs>0) & (nBasalSpcs>0)


plot(table(core=nCoreSpcs,basal=nBasalSpcs,
           ifelse(nOutSpcs>0,"out","noOut")),
     col=rainbow(5), xlab="Number of core species", 
     ylab="Number of basal species",
     main=paste0(sum(nOutSpcs==0)," noOut (trees missing outgroup), ",
                 sum(nOutSpcs>0)," out (trees with outgroup) "))

```


### Check if trees have the in-group/out-group split and core split

Limit the test to only those trees that have atleast one ortholog in the basal, core and out group (`r sum(hasOrthoInEach)` trees)

A well formed tree would have the out-group as a single clan and the core group as a single clan.


```{r inOutSplit, echo=FALSE}


# Check if tips (given by logical vector isInGrp) form a clan in the tree
isClan <- function(tree, isInGrp){
  clans = getClans(tree)
  any(apply(clans,1,function(clan){
      all(clan==isInGrp)
  }))
}


sapply(allTrees,function(tree){
  isClan(tree,grepl("(Os|Zm|Sb)_R",tree$tip.label))
}) -> isOutGroupClan

sapply(allTrees,function(tree){
  isClan(tree,grepl("^(H|B)",tree$tip.label))
}) -> isCoreClan

table(isOutGroupClan = isOutGroupClan[hasOrthoInEach],isCoreClan = isCoreClan[hasOrthoInEach])

```

### Use clanFinder algorithm to split big groups

In some situations orthoMCL has grouped together paralogs that were duplicated before the out-groups split from the pooids (out-paralogs). In that case the tree would not have a single out-group clan but several. ClanFinder seperates those trees into clans that each contain single out-group clan (and minimum 3 orthologs in the in-group).

Trees that don't have a valid outgroup split will be attempted to split up using clanFinder.

```{r defClanFinder, echo=FALSE}
# Find all clans in the tree that contain exactly one in-clan and one out-clan
clanFinder = function(tre, 
                      minSize=3, 
                      ut=c("Os_R", "Sb_R", "Zm_R"),
                      output=c("tree", "tips"),
                      summaryFUN=NULL) { 
  
    clans = getClans(tre) # phangorn
    
    # get species names (split on pipe)
    specs = sapply(strsplit(colnames(clans), "|", fixed=T), "[", 1)

    # convert clans to other format
    clanList_all = apply(clans, 1, function(r) which(as.logical(r)))
    
    # which clans >= minSize and does not contain any outgroup species?
    clanList_in = clanList_all[ vapply(clanList_all, function(cln){
        length(cln) >= minSize  &&  !any(specs[cln] %in% ut)
      }, logical(1))]

    # which clans >= 2 and only contain outgroup species?
    clanList_ut = clanList_all[vapply(clanList_all, function(cln) {
        length(cln) >= 2 && all(specs[cln] %in% ut)
      }, logical(1))]

    # for each clan
    #   clan is goodClan if: clan > minSize and 
    #   contains exactly one in-clan (size >= minsize) and one out-clan
    goodClan = vapply(clanList_all, function(cln) {
        if(length(cln) < minSize+1) return(F)
        sp = specs[cln]             # associated species
        utgruppe = cln[sp %in% ut]  # out group clan members
        ingruppe = cln[!sp %in% ut] # in group clan members
        if(length(ingruppe) < minSize || length(utgruppe) == 0) # at least 'minsize' in and 1 out
            return(F)
        in_isClan = list(ingruppe) %in% clanList_in  # test if in group members form a clan
        ut_isClan = length(utgruppe) == 1 || list(utgruppe) %in% clanList_ut # same for out members (if more than 1)
        in_isClan && ut_isClan
    }, logical(1))
    
    tip_seq = seq_along(tre$tip.label)
    switch(match.arg(output),
        tree = lapply(clanList_all[goodClan], function(cln) drop.tip(tre, setdiff(tip_seq, cln))),
        tips = lapply(clanList_all[goodClan], function(cln) colnames(clans)[cln]))
}

```


```{r runClanFinder, echo=FALSE, cache=TRUE}

# The trees that don't already have a valid outgroup split and have atleast one outgroup ortho
# and at least 3 ingroup orthos
table( hasOutAnd3In = (nOutSpcs>0) & (nCoreSpcs+nBasalSpcs >= 3),isOutGroupClan)

mustBeSplit <- !isOutGroupClan & (nOutSpcs>0) & (nCoreSpcs+nBasalSpcs >= 3)

splitTrees <- lapply(allTrees[mustBeSplit], clanFinder, output="tree")


# Check out if any clans are overlapping
# any(sapply(splitTrees,function(trees){
#   any(table(unlist(lapply(trees,"[[","tip.label")))>1)
# }))

plot(table(sapply(splitTrees,length)),xlab="Number of good clans found",
     ylab="Number of trees",main="Clans found per tree:")

# rename the split trees
lapply(splitTrees,function(trees){
  setNames(trees,seq_along(trees))
}) -> renamedTrees

# unlist the split trees
treesSplit <- unlist(renamedTrees,recursive = F)

```

ClanFinder found `r length(treesSplit)` good clans in `r length(splitTrees)` trees.

```{r summarizeTreesPlot, echo=FALSE, cache=TRUE}

getLeaves <- function(node, tree){
  if( node <= length(tree$tip.label)){
    return(node)
  } else {
    return( unlist(lapply( tree$edge[tree$edge[,1]==node,2], getLeaves, tree)) )
  }
}


summarizeTrees <- function(trees, main){
  # remove all but the species label
  lapply(trees,function(tree){
    tree$tip.label <- stringr::str_extract(tree$tip.label,"^[^\\|]*")
    return(tree)
  }) -> ttrees
  
  
  
  nodeValidClans <- integer(0)
  nodeValidClans2 <- integer(0)
  nodeHasAllTips <- integer(0)
  # for each internal node in the guide tree
  for( node in 12:max(myTree$edge)){
    tips <- myTree$tip.label[getLeaves(node, myTree)]
    
    sapply(ttrees,function(tree){
      all(tips %in% tree$tip.label)
    }) -> hasAllTips
    
    sapply(ttrees,function(tree){
      isClan(tree, tree$tip.label %in% tips)
    }) -> validClans
    
    nodeValidClans <- c(nodeValidClans,sum(validClans & hasAllTips))
    nodeValidClans2 <- c(nodeValidClans2,sum(validClans))
    nodeHasAllTips <- c(nodeHasAllTips,sum(hasAllTips))
  }
  
  # get tip counts
  tipClans <- integer(0)
  tipCounts <- NULL
  # for each tip
  for( tip in myTree$tip.label){
    sapply(ttrees,function(tree){
      sum(tree$tip.label == tip)
    }) -> nTips
    
    sapply(ttrees,function(tree){
      isClan(tree, tree$tip.label == tip)
    }) -> validClans
    
    tipClans <- c(tipClans,sum(nTips>1 & validClans))
    tipCounts <- rbind(tipCounts, c(sum(nTips==0),
                                    sum(nTips==1),
                                    sum(nTips>1 & validClans),
                                    sum(nTips>1 & !validClans)))
  }
  
  
  # root node always has valid clan
  nodeValidClans2[1] <- length(ttrees)
  nodeValidClans[1] <- nodeHasAllTips[1]
  
  cbind(
    c01 = nodeValidClans2-nodeValidClans,
    c11 = nodeValidClans,
    c10 = nodeHasAllTips-nodeValidClans,
    c00 = length(ttrees)-(nodeValidClans2+nodeHasAllTips-nodeValidClans)
  )/length(ttrees) -> tipsAndClansPie
  
  c(
    c01 = "#A0A0FF",
    c11 = "#A060A0",
    c10 = "#FFA0A0",
    c00 = "#FFFFFF"
  ) -> tipsAndClansPieCol
  
  
  treePlot(myTree, main=main)
  nodelabels(pie=tipsAndClansPie,piecol = tipsAndClansPieCol)
  
  tipPieCol <-  c("#FFFFFF", "#BAE4B3", "#74C476", "orange")
  tiplabels(pie = tipCounts/length(ttrees), adj=0.25,
            piecol = tipPieCol)
  
  legend("bottomleft",legend = c("proper clade","has all tips",
                                 "has this tip","in-paralog",
                                 "out-paralog"),
         fill = c(tipsAndClansPieCol[c(1,3)],tipPieCol[-1]))
  
}

summarizeTrees(treesSplit, 
               main=paste0("Trees found by clanFinder (",
                           length(treesSplit),")"))
summarizeTrees(allTrees[isOutGroupClan], 
               main=paste0("Trees already having out-group (",
                           sum(isOutGroupClan),")"))

```

### Tree quality

Minimum requirement for a tree:

* Have at least one out-group species
* Have outgroup split
* have at least one basal and one of the de-novo core species


```{r minimumReq, echo=FALSE}

# use the split trees and the already good trees
trees <- c(treesSplit,allTrees[isOutGroupClan])


sapply(trees, function(tree){
  sum(grepl("^(HoVu|BrDi)",tree$tip.label))
}) -> nCoreDeNovoSpcs

sapply(trees, function(tree){
  sum(grepl("^(MeNu|NaSt|StLa)",tree$tip.label))
}) -> nBasalSpcs2

# minimum requirement is
# * have outgroup and outgroup split
# * have at least 1 basal and 1 de-novo core species
nNoGoodTrees <- sum(!isOutGroupClan & !mustBeSplit) + 
  sum(sapply(splitTrees,length)==0) +
  sum(nCoreDeNovoSpcs==0 | nBasalSpcs2==0)

nMinimumReqTrees <- sum(nCoreDeNovoSpcs>0 & nBasalSpcs2>0)

```

The number of trees that pass the minimum requirements: `r nMinimumReqTrees`

The number of trees that was discarded: `r nNoGoodTrees`

Out of the trees that pass the minimum requirements there are several metrics that can be used to classify the trustworthiness of the trees.

* Number of species represented: Minimum req. < 2 basal and 2 core < All de-novo < All species
* The existance of paralogs
* The core species form a clan
* Each species form a separate clan (the pairs MeNu1/MeNu2, HoVu/Hv_R and BrDi/Bd_R are treated as single species)

```{r qualityBreakdownPlot, echo=FALSE}

# continue calculation on trees that meet the minimum requirements
trees <- trees[nCoreDeNovoSpcs>0 & nBasalSpcs2>0]

# Calculate further requirements for tree quality

# species
spcs <- c("MeNu1","MeNu2","NaSt","StLa","HoVu","BrDi","Hv_R","Bd_R")
sapply(spcs, function(spc){
  sapply(trees, function(tree){
    sum(grepl(paste0("^",spc),tree$tip.label))
  })
})-> nSpcs2

nCoreSpc <- rowSums(nSpcs2[,c("HoVu","BrDi","Hv_R","Bd_R")]>0)
nCoreDeNovoSpc <- rowSums(nSpcs2[,c("HoVu","BrDi")]>0)
nBasalSpc <- rowSums(nSpcs2[,c("MeNu1","MeNu2","NaSt","StLa")]>0)

has2BasalAnd2Core <- nCoreSpc>1 & nBasalSpc>1
hasAllDenovo <- nCoreDeNovoSpc==2 & nBasalSpc==4
hasAllPooids <- nCoreSpc==4 & nBasalSpc==4

# hasSpcs <- factor(ifelse(!has2BasalAnd2Core,"has1BasalAnd1CoreDeNovo",
#               ifelse(!hasAllDenovo,"has2BasalAnd2Core",
#                      ifelse(!hasAllPooids,"hasAllDenovo","hasAll"))))

hasSpcs <- factor(ifelse(!has2BasalAnd2Core,"*",
              ifelse(!hasAllDenovo,"**",
                     ifelse(!hasAllPooids,"***","****"))))

# has core split
sapply(trees,function(tree){
  isClan(tree,grepl("^(H|B)",tree$tip.label))
}) -> isCoreClan2


hasParalogs <- rowSums(nSpcs2>1)>0

sapply(c("H","B","MeNu","NaSt","StLa"), function(spc){
  sapply(trees,function(tree){
    isInGrp <- grepl(paste0("^",spc),tree$tip.label)
    if( sum(isInGrp) < 2){
      return(TRUE) # return TRUE if tip not existing
    } else {
      return(isClan(tree,isInGrp))      
    }
  })
}) -> isSpcClan

allSpcAreClans <- apply(isSpcClan,1,all)

plot(table(allSpcAreClans, 
           hasParalogs, 
           ifelse(isCoreClan2,"coreClan","notClan"), 
           hasSpcs),
     col=rainbow(4), 
     main="Breakdown of tree quality based on various metrics")

legend( "bottom",
        inset = c(0, -0.2), bty = "n", 
        x.intersp=0.5, xjust=0,yjust=0, 
        legend=c("Minimum req.", "2 basal and 2 core",
                 "All de-novo", "All species"),
        fill=rainbow(4), cex = 0.75, xpd = TRUE, horiz=T)

```

## Paralog Heatmap

```{r paralogHeatmap, echo=FALSE}
writeLines(text = paste0(names(trees),": ", 
                         lapply(trees, function(tree){
                           paste(tree$tip.label, collapse=" ")
                           })),
             con = "~/temp/treeGrps.txt")

treeGrps <- loadOrthoGrpsArray(orthoGrpFile = "~/temp/treeGrps.txt")

treeLens <- unlist(lapply(treeGrps,length))
dim(treeLens) <- dim(treeGrps)
dimnames(treeLens) <- dimnames(treeGrps)

idx <- allSpcAreClans & hasParalogs & isCoreClan2 & hasAllDenovo

heatmap(treeLens[idx,!grepl("_R",colnames(treeLens))],
        scale="none",labRow = "", 
        breaks=c(0:3,2^(2:5))-0.5,
        col=RColorBrewer::brewer.pal(n = 7,"Blues"))

```

```{r testCode, echo=FALSE,eval=FALSE}

treePlotU2 <- function(tree, ...){
  plot.phylo(tree,show.tip.label = F,type="u",...)
  tiplabels(pch=20,col=ifelse(grepl("(Os|Zm|Sb)_R",tree$tip.label),"red","blue"))
}

idx <- 1208
idx <- 1247
idx <- idx + 1
tree <- allTrees[[idx]]
clans <- clanFinder(tree, output = "tree")

treePlotU(tree)
for( i in seq_along(clans)){
  tiplabels( pch=ifelse(tree$tip.label %in% clans[[i]]$tip.label,i,NA),cex=2)
}
legend("bottomright",pch=c(20,20,20,1,2),col=c("red","blue","green","black","black"),legend=c("outgroup","ingroup (basal)","ingroup (core)",paste("clan",1:2)),pt.cex=c(1,1,1,2,2))
```

